---
title: "Testing Rmarkdown"
author: "luisa.barbanti@hslu.ch"
date: "October 4th, 2023"
output: word_document
---

We create a vector that is the sequence of one to 20 and display it.

```{r}
x <- 1:20
x
```

Now we can compute the mean _of the vector x_.

```{r}
mean(x)
```

Now we want to plot the vector x **against another vector y**. This latter is a vector of 20 random numbers.

```{r}
y <- rnorm(n = 20)
plot(y ~ x)
```

Note that you can compile pdf documents, as long as LaTeX is installed on you machine. To quickly install LaTeX on your machine run the following commands.

```{r eval=FALSE}
install.packages("tinytex")
library(tinytex)
install_tinytex()
```

Then you can write things like this: $y = x + \epsilon$.

This is a (non-exhaustive) list of things you can do with Rmarkdown:

* name chunks
* use "engines" other than R (e.g. SQL or python)
* set chunk arguments such as:
    + ``echo = FALSE`` (i.e. R-code is not reported)
    + ``results = "hide"`` (R-output is not reported)
    + ``fig.keep = "none"`` (plots are not reported)
    + ``message = FALSE`` (messages such as package loading messages are not reported)
* use LaTeX commands (e.g. \\bigskip to add vertical space)
* or use html commands (e.g. \<br> to create a line break)
* ... very many things


<!-- Finally, this is a comment and won't be compiled -->

