---
title: "Appendix: Additional Models & Explorations"
author: "Dongyuan Gao, Ramiro, Cyriel"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Re-load necessary libraries and data setup from main report logic
library(tidyverse)
library(caret)
library(mgcv)
library(pROC)
library(gridExtra)

# Basic Data Loading (Simplified for standalone running)
# Note: Ensure paths match your project structure
if(file.exists('data/data_choosing_process/Cyriel/data.csv')) {
  df <- read.csv2('data/data_choosing_process/Cyriel/data.csv', fileEncoding = 'UTF-8-BOM')
  
  numeric_cols <- c('Admission.grade', 'Previous.qualification..grade.',
                    'Age.at.enrollment', 'Mother.s.qualification',
                    'Father.s.qualification', 'Curricular.units.1st.sem..grade.',
                    'Curricular.units.2nd.sem..grade.', 'GDP',
                    'Unemployment.rate', 'Inflation.rate')

  for (col in numeric_cols) {
    if (col %in% names(df)) {
      df[[col]] <- as.numeric(as.character(df[[col]]))
    }
  }
}
```

# A.1 Binomial GLM: Predicting High vs Low Admission Scores

*Originally from Section 5.2 of the main report.*

This model attempts to classify students into "High" or "Low" admission categories based on their background. While interesting for understanding entry profiles, it is less critical for the dropout prediction task.

```{r binomial-admission-model}
median_admission <- median(df$Admission.grade, na.rm = TRUE)
df$High_Admission <- ifelse(df$Admission.grade >= median_admission, 1, 0)

features_admission <- c(
  'Previous.qualification..grade.',
  'Age.at.enrollment',
  'Mother.s.qualification',
  'Father.s.qualification',
  'Scholarship.holder',
  'Gender',
  'Debtor',
  'Tuition.fees.up.to.date',
  'Curricular.units.1st.sem..approved.',
  'Curricular.units.1st.sem..grade.',
  'Unemployment.rate',
  'Inflation.rate',
  'GDP'
)

df_admission <- df %>%
  select(all_of(c(features_admission, 'High_Admission'))) %>%
  mutate(across(-High_Admission, as.numeric)) %>%
  na.omit()

set.seed(42)
train_index_adm <- createDataPartition(df_admission$High_Admission, p = 0.8, list = FALSE)
train_adm <- df_admission[train_index_adm, ]
test_adm  <- df_admission[-train_index_adm, ]

# Model Training
glm_adm <- glm(High_Admission ~ ., data = train_adm, family = binomial(link = "logit"))
summary(glm_adm)

# Evaluation
pred_prob_adm <- predict(glm_adm, test_adm, type = "response")
pred_class_adm <- ifelse(pred_prob_adm > 0.5, 1, 0)
confusionMatrix(factor(pred_class_adm), factor(test_adm$High_Admission))
```


# A.2 Poisson GLM: 2nd Semester Course Approvals

*Originally from Section 6.3 of the main report.*

This model predicts the count of approved courses in the 2nd semester. It was removed from the main report because its findings (financial stability and 1st semester success are key) overlap heavily with the Linear Regression and 1st Semester Poisson models.

```{r poisson-2nd-sem}
TARGET_2 <- 'Curricular.units.2nd.sem..approved.'
FEATURES_2 <- c(
    'Previous.qualification..grade.',
    'Admission.grade',
    'Age.at.enrollment',
    'Scholarship.holder',
    'Debtor',
    'Tuition.fees.up.to.date',
    'Curricular.units.1st.sem..approved.',
    'Curricular.units.1st.sem..grade.',
    'Curricular.units.2nd.sem..enrolled.',
    'Unemployment.rate',
    'GDP'
)

df_model2 <- df %>%
    select(all_of(c(FEATURES_2, TARGET_2))) %>%
    mutate(across(everything(), ~ as.numeric(as.character(.)))) %>%
    drop_na()

set.seed(42)
trainIndex2 <- createDataPartition(df_model2[[TARGET_2]], p = 0.8, list = FALSE)
train_data2 <- df_model2[trainIndex2, ]
test_data2  <- df_model2[-trainIndex2, ]

poisson_model2 <- glm(as.formula(paste(TARGET_2, "~", paste(FEATURES_2, collapse = " + "))), 
                      data = train_data2, family = poisson(link = "log"))

# Prediction & Metrics
y_pred2 <- predict(poisson_model2, newdata = test_data2, type = "response")
rmse2 <- sqrt(mean((test_data2[[TARGET_2]] - y_pred2)^2))
mae2 <- mean(abs(test_data2[[TARGET_2]] - y_pred2))
cat(sprintf("Test RMSE: %.4f | MAE: %.4f", rmse2, mae2))
```

# A.3 Generalized Additive Models (GAM)

*Originally from Section 7 of the main report.*

GAMs were used to explore non-linear relationships. They confirmed that most relationships (e.g., grades) are largely linear, but some variables (like Age) show non-linear patterns.

## A.3.1 Gaussian GAM (Grades)
```{r gam-grade}
# Using limited features for demonstration
gam_grade <- gam(Curricular.units.2nd.sem..grade. ~ 
                   s(Admission.grade) + s(Age.at.enrollment) + factor(Scholarship.holder),
                 data = df, method = "REML")
plot(gam_grade, pages = 1)
summary(gam_grade)
```

## A.3.2 Poisson GAM (Course Counts)
```{r gam-count}
gam_count <- gam(Curricular.units.2nd.sem..approved. ~ 
                   s(Admission.grade) + s(Age.at.enrollment) + factor(Scholarship.holder),
                 data = df, family = poisson, method = "REML")
summary(gam_count)
```
