---
title: "Predicting Student Success:<br>A Multi-Model Machine Learning Approach"
subtitle: "-- Early Warning System for Student Dropout Risk"
author: "Dongyuan Gao, Ramiro, Cyriel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: cosmo
    df_print: paged
---

```{css custom-styles, echo=FALSE}
h1, h2, h3, h4, h5, h6 {
  color: #111111;
  font-weight: 700;
}
h1 {
  font-size: 2.3rem;
}
h2 {
  font-size: 2.1rem;
}
h3 {
  font-size: 1.8rem;
}
h4 {
  font-size: 1.5rem;
}
.caption, caption {
  color: #111111;
  font-weight: 600;
}
body {
  color: #222222;
  font-size: 1.8rem;
  font-weight: 470;
}
/* Compact chunk toggle buttons */
.btn.btn-default.btn-xs.code-folding-btn {
  padding: 2px 6px;
  font-size: 10px;
  line-height: 1.1;
  margin: 2px 0 4px 0;
}
/* Title block */
.title {
  font-size: 2.4rem;
}
.subtitle {
  font-size: 2.3rem;
}
.author, .date {
  font-size: 2.1rem;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

# ============================================================================
# PACKAGE INSTALLATION 
# Uncomment and run once these lines if you need to install the packages
# ============================================================================

# --- Core Data Manipulation & Visualization ---
# install.packages("tidyverse")    # Data manipulation (dplyr, tidyr, etc.) and ggplot2
# install.packages("ggplot2")      # Grammar of graphics plotting (included in tidyverse)
# install.packages("scales")       # Scale functions for visualization
# install.packages("gridExtra")    # Arrange multiple plots in a grid
# install.packages("corrplot")     # Correlation matrix visualization

# --- Machine Learning Utilities ---
# install.packages("caret")        # Classification and Regression Training
# install.packages("broom")        # Tidy model outputs

# --- Generalized Linear Models & GAM ---
# install.packages("mgcv")         # Generalized Additive Models (GAM)
# install.packages("gratia")       # GAM visualization and diagnostics
# install.packages("pROC")         # ROC curve analysis

# --- Neural Networks ---
# install.packages("nnet")         # Single hidden layer neural network (Lab 1 style)
# install.packages("neuralnet")    # Multi-layer neural network (Lab 2 style)
# install.packages("ROCR")         # ROC curves for neural networks
# install.packages("NeuralNetTools") # Visualize neural network structure

# --- Support Vector Machines ---
# install.packages("e1071")        # SVM implementation (svm function)
# install.packages("kernlab")      # SVM kernels

# ============================================================================

# Load required libraries
library(tidyverse)
library(ggplot2)
library(e1071)         # SVM
library(caret)         # ML utilities
library(kernlab)       # SVM kernels
library(mgcv)          # GAM
library(pROC)          # ROC curves
library(gridExtra)     # Multiple plots
library(corrplot)      # Correlation plots
library(broom)         # Tidy model outputs
library(scales)        # Scaling functions
library(gratia)        # GAM visualization
library(nnet)          # Neural Network (Lab 1 style)
library(neuralnet)     # Neural Network (Lab 2 style)
library(ROCR)          # ROC curves for neural networks
library(NeuralNetTools) # Visualize network structure

options(warn = -1)
```

# Summary

**Our Storyline:** Higher education institutions face significant challenges with **student drop-outs**. A fictional university has engaged our education consulting firm to understand what influences student dropout and how to better prevent dropouts. The university wants to **understand and improve academic performance**, and design **interventions** that reduce failed academic lives. Understanding which students are **at risk of dropping out**, and timely support that can change academic and life trajectories.

**Our Data:** We have chosen a real case dataset from **University of California, Irvine (UCI)**, to simulate the fictional case. We found the dataset because the data repository website appeared once in our Machine Learning 1 lecture.

**Our Approach:** We analyzed **4,424 students** from a Portuguese higher education institution using **six** machine learning models to predict student outcomes (**Dropout**, **Enrolled**, **Graduate**) and **understand the factors that drive academic success**.
**Project Repository:** [Click here to view full code & repositorz on GitHub](https://github.com/dydy2010/machine_learning_I_Project)

## Key Findings:

-   **First & Second semester performance is the strongest predictor** of academic outcomes across all models.
-   **2 Early warning systems** could be built upon our models.
-   **Non-linear relationships exist** between age, course load, and success.
-   **Model accuracy ranges from 70-90%** depending on the prediction task, with graduation prediction achieving the highest discrimination (**AUC ≈ 0.87**)

**Business Impact:** These models enable our client university to **identify at-risk students early**, **allocate resources**, and implement **targeted interventions** that can improve graduation rates.

------------------------------------------------------------------------

# 1. Introduction & Data Context

## 1.1 Business Questions & Model Mapping

**Student dropout** is a critical issue in higher education. As noted by the OECD, "In Switzerland, STEM completion rates are 68%, lower than those for health and welfare, at 81%" (OECD, 2025). To address it, we deploy six modeling approaches, each answering a specific business question:

1.  **How do 1st-semester results predict students' final graduation/dropout?** (Binomial GLM / GAM)
    *   *Purpose:* Early intervention based on initial performance.
2.  **Can we predict 2nd-semester grades?** (Linear Regression / Gaussian GAM)
    *   *Purpose:* Forecasting academic performance. The prediction result could be beneficial for intervention methods.
3.  **Can we predict how many courses will a student pass in the 1st semester?** (Poisson GLM)
    *   *Purpose:* Identifying students struggling with studies. The prediction result could be beneficial for intervention methods.
4.  **How do major factors like 2nd semester performance and finance predict students' final graduation/dropout?** (SVM / Neural Net)
    *   *Purpose:* High-accuracy classification for targeted support.

By predicting **dropout risk early**, institutions can intervene with:

-   Academic support programs
-   Financial aid counseling
-   Mental health services
-   Personalized study plans

## 1.2 Team Contributions

This project was completed collaboratively by three team members, with each member taking the lead on specific models:

| Team Member | Models Led | Contributions |
|-------------|------------|---------------|
| **Cyriel**| Linear Regression, GLM (Binomial & Poisson), GAM | Exploratory Analysis |
| **Ramiro** | Support Vector Machines | Data Preprocessing |
| **Dongyuan Gao** | Neural Networks | Cross-validation, <br> Models proof-reading & Report Editing |


## 1.3 Data Source and Selection Process

### Choosing Our Dataset

**Our goal** for this project is to deliver a **solid report** while using it as an opportunity to **enhance our understanding** of **machine learning concepts** and practices. With this in mind, we explored several *potential datasets* for our final analysis.

We considered **three main dataset options**. First, the Social Media Fact Sheet from PEW Research Center, which would have allowed us to analyze social media usage patterns. Second, medical quality service data in Switzerland from the Federal Office of Public Health (available through opendata.swiss). Third, student data for predicting dropout and academic success from the UC Irvine Machine Learning Repository.

After careful consideration, we chose the third option from the UC Irvine repository. The **data structure and scope align well with our goal** of getting familiarized with machine learning concepts and practices and producing a solid report as our first attempt in the ML world. Though the other two datasets were interesting to explore, their structure, diversity, and scope did not suit our *learning objectives* as well as the student success dataset. For the exploration experience, we didn't read any papers related to the dataset online while doing the project.

### Dataset Characteristics

**Source:** UCI Machine Learning Repository - Predict Students' Dropout and Academic Success\
**Institution:** Portuguese higher education institution\
**Time Period:** Multiple academic years\
**Sample Size:** 4,424 students with 37 variables (e.g., age, 1st-semester grades, credits completed)

The dataset provides a comprehensive view of student characteristics and outcomes. It is well-suited for applying multiple machine learning models.

**Target Variable (3 factors):**

-   **Dropout:** Students who left the program (1,421 students, 32%)
-   **Enrolled:** Students currently continuing (794 students, 18%), in some models counted as binary 'not-dropout'.
-   **Graduate:** Students who completed successfully (2,209 students, 50%)

The distribution shows a balanced dataset with a slight majority of graduates. It reflects the institution's overall success rate and provides sufficient samples across all outcome categories for model training.

```{r load-data, include=FALSE}
# Load data
df <- read.csv2('data/data_choosing_process/Cyriel/data.csv',
                fileEncoding = 'UTF-8-BOM')

# Convert numeric columns from character to numeric
numeric_cols <- c('Admission.grade', 'Previous.qualification..grade.',
                  'Age.at.enrollment', 'Mother.s.qualification',
                  'Father.s.qualification', 'Curricular.units.1st.sem..grade.',
                  'Curricular.units.2nd.sem..grade.', 'GDP',
                  'Unemployment.rate', 'Inflation.rate')

for (col in numeric_cols) {
  if (col %in% names(df)) {
    df[[col]] <- as.numeric(as.character(df[[col]]))
  }
}
```
## 1.4 Feature Categories and detailed variables

Our dataset includes **36 predictors** in the following categories: Demographic Information, Socioeconomic Factors, Academic Background, Academic Performance, Financial Indicators, and Macroeconomic Context.
To make the business case and model results easier to follow, the table below summarizes the most important variables used.

```{r variable-overview-table, results="asis", echo=FALSE}
variable_overview <- data.frame(
  Variable = c(
    "Target",
    "Admission.grade",
    "Previous.qualification..grade.",
    "Age.at.enrollment",
    "Gender",
    "Scholarship.holder",
    "Debtor",
    "Tuition.fees.up.to.date",
    "Curricular.units.1st.sem..enrolled.",
    "Curricular.units.1st.sem..approved.",
    "Curricular.units.1st.sem..grade."
  ),
  Meaning = c(
    "Student outcome (Dropout / Enrolled / Graduate)",
    "Admission exam/entry score used for program entry",
    "Grade from the previous qualification (prior education)",
    "Student age when enrolling in the program",
    "Student gender (coded numerically in the dataset)",
    "Whether the student holds a scholarship (0/1)",
    "Whether the student is classified as a debtor (0/1)",
    "Whether tuition fees are paid up to date (0/1)",
    "Number of courses the student enrolled in (1st semester)",
    "Number of courses successfully passed (1st semester)",
    "Average grade in the 1st semester (0–20 scale)"
  ),
  Type = c(
    "Categorical (3 classes)",
    "Numeric",
    "Numeric",
    "Numeric",
    "Categorical (coded)",
    "Binary (0/1)",
    "Binary (0/1)",
    "Binary (0/1)",
    "Count",
    "Count",
    "Numeric"
  ),
  stringsAsFactors = FALSE
)
rownames(variable_overview) <- NULL

# Use kableExtra if available; otherwise fall back to plain kable
if (requireNamespace("kableExtra", quietly = TRUE)) {
  knitr::kable(
    variable_overview,
    caption = "Key dataset variables (collapsed view)"
  ) %>%
    kableExtra::kable_styling(font_size = 10, bootstrap_options = c("striped")) %>%
    kableExtra::column_spec(1, bold = TRUE, width = "12em") %>%
    kableExtra::column_spec(2, width = "18em") %>%
    kableExtra::column_spec(3, width = "10em")
} else {
  knitr::kable(
    variable_overview,
    caption = "Key dataset variables (collapsed view)",
    align = c("l", "l", "c")
  )
}
```


### Why This Dataset Works for Our Analysis

The dataset is suitable for applying all required machine learning methods:

**For Linear Regression:** We have **continuous grade variables (0-20 scale)**, allowing us to predict second semester performance based on first semester results and student characteristics.

**For Binomial GLM:** The three-class target can be converted to **Graduate vs Dropout**, enabling logistic regression with **interpretable odds ratios**.

**For Poisson GLM:** **Count variables** are present (approved courses, evaluations, failed courses), making it ideal for count-based modeling.

**For GAM, Neural Networks, and SVM:** The mix of **continuous and categorical predictors** plus a **classification task** lets us explore **non-linear relationships** and **complex interactions**.

The dataset contains **4,424 observations**—large **enough** for robust train-test splits and complex **training**, yet manageable. With **36 predictors** across demographic, academic, financial, and socioeconomic domains, we have the complexity to deliver **insights for our client**.

------------------------------------------------------------------------

# 2. Exploratory Data Analysis

## 2.1 Target Variable Distribution

```{r target-distribution, fig.height=5}
# Create a bar plot with percentages
target_summary <- df %>%
  group_by(Target) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

ggplot(target_summary, aes(x = Target, y = count, fill = Target)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(aes(label = paste0(round(percentage, 1), "%\n(n=", count, ")")),
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Dropout" = "#E74C3C",
                                "Enrolled" = "#F39C12",
                                "Graduate" = "#27AE60")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Distribution of Student Outcomes",
       x = "Outcome", y = "Number of Students", fill = "Outcome") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

**Key Observation:** The dataset shows a **reasonable distribution** for modeling, with 50% graduates, 32% dropouts, and 18% still enrolled. This balance allows models to learn patterns for all outcomes effectively. Depending on the model and task, we may need to **adjust** the classes to **binary(dropout or not dropout)**.

## 2.2 First Semester Performance Analysis

```{r first-sem-analysis, fig.height=5}
# Prepare data for first semester grades
first_sem_data <- df %>%
  filter(!is.na(Curricular.units.1st.sem..grade.) &
         Curricular.units.1st.sem..grade. > 0) %>%
  select(Target, Curricular.units.1st.sem..grade.,
         Curricular.units.1st.sem..approved.)

# Box plot for first semester grades
p1 <- ggplot(first_sem_data, aes(x = Target, y = Curricular.units.1st.sem..grade., fill = Target)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Dropout" = "#E74C3C",
                                "Enrolled" = "#F39C12",
                                "Graduate" = "#27AE60")) +
  labs(title = "First Semester Grades by Outcome",
       x = "Outcome", y = "First Semester Grade (0-20)") +
  theme(legend.position = "none")

# Box plot for courses approved
p2 <- ggplot(first_sem_data, aes(x = Target, y = Curricular.units.1st.sem..approved., fill = Target)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("Dropout" = "#E74C3C",
                                "Enrolled" = "#F39C12",
                                "Graduate" = "#27AE60")) +
  labs(title = "Courses Approved in First Semester",
       x = "Outcome", y = "Number of Courses Approved") +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```

**Insight:** *First semester performance* shows dramatic differences across outcomes. Graduate students average **13+ grade points** and complete most enrolled courses(approved), while dropout students struggle with grades around **9-10** and complete fewer courses. This suggests that *first semester performance* *could be a critical indicator* for dropout prediction.

## 2.3 Financial & Socioeconomic Factors

```{r financial-factors, fig.height=5}
# Prepare financial data
financial_data <- df %>%
  select(Target, Tuition.fees.up.to.date, Scholarship.holder, Debtor) %>%
  mutate(Tuition.fees.up.to.date = factor(Tuition.fees.up.to.date,
                                           levels = c(0, 1),
                                           labels = c("Not Current", "Current")),
         Scholarship.holder = factor(Scholarship.holder,
                                     levels = c(0, 1),
                                     labels = c("No", "Yes")),
         Debtor = factor(Debtor,
                        levels = c(0, 1),
                        labels = c("No", "Yes")))

# Tuition status
p1 <- ggplot(financial_data, aes(x = Tuition.fees.up.to.date, fill = Target)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Dropout" = "#E74C3C",
                                "Enrolled" = "#F39C12",
                                "Graduate" = "#27AE60")) +
  labs(title = "Tuition Payment Status",
       x = "Tuition Fees Status", y = "Proportion", fill = "Outcome")

# Scholarship status
p2 <- ggplot(financial_data, aes(x = Scholarship.holder, fill = Target)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Dropout" = "#E74C3C",
                                "Enrolled" = "#F39C12",
                                "Graduate" = "#27AE60")) +
  labs(title = "Scholarship Holder Status",
       x = "Has Scholarship", y = "Proportion", fill = "Outcome")

grid.arrange(p1, p2, ncol = 2)
```

**Insight:** Students that currently **pay tuition on time**(tuition.fees.up.to.date) show significantly higher graduation rates. **Scholarship** holders also demonstrate better outcomes, suggesting that financial situation could play an important role in student success.

## 2.4 Age Distribution

```{r age-distribution, fig.height=4}
ggplot(df, aes(x = Age.at.enrollment, fill = Target)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Dropout" = "#E74C3C",
                                "Enrolled" = "#F39C12",
                                "Graduate" = "#27AE60")) +
  labs(title = "Age Distribution by Student Outcome",
       x = "Age at Enrollment", y = "Density", fill = "Outcome") +
  theme(legend.position = "bottom")
```

**Insight:** Age shows interesting patterns: common students age **(18-22)** have **higher graduation rates**, while both younger and significantly older students face elevated dropout risks. This pattern suggests that age effects may not follow a linear trend, which we could explore using models that can capture such **non-linear** relationships (e.g., *GAMs* with smooth terms, or GLMs with *polynomial/spline* features).

------------------------------------------------------------------------

# 3. Modeling Approach & Strategy

## 3.1 Model Characteristics

We employ six modeling techniques, each with distinct strengths:

| Model                 | Purpose                     | Strengths                                           |
|------------------|----------------------|--------------------------------|
| **Linear Regression** | Predict continuous grades   | Interpretable, fast, baseline for comparison        |
| **GLM (Binomial)**    | Binary classification       | Probabilistic output, interpretable odds ratios     |
| **GLM (Poisson)**     | Count data prediction       | Appropriate for non-negative integer outcomes       |
| **GAM**               | Non-linear relationships    | Captures curves, visualizable, flexible             |
| **Neural Networks**   | Complex patterns            | Learns interactions automatically, high capacity    |
| **SVM**               | Robust classification       | Maximum margin, kernel tricks, outlier resistant    |

## 3.2 Research Questions

Each model addresses specific questions:

1.  **Linear Regression:** How do student characteristics linearly predict **second semester grades**?
2.  **GLM (Binomial):** What factors distinguish **graduates from dropouts** in probabilistic terms?
3.  **GLM (Poisson):** What predicts the **count of approved courses** in each semester?
4.  **GAM:** Are there **non-linear relationships** that traditional linear models miss?
5.  **Neural Networks:** Can **deep learning** uncover **hidden interaction patterns**?
6.  **SVM:** Can we build a **robust classifier** that maximizes the margin between success and failure?

## 3.3 Evaluation Strategy

-   **Training/Test Split:** **80/20 stratified split** to maintain class balance
-   **Regression Metrics:** **R²**, **RMSE**, **MAE**
-   **Classification Metrics:** **Accuracy**, **AUC-ROC**, **Confusion Matrix**
-   **Cross-Model Comparison:** Identify **consistent predictors** across methods

------------------------------------------------------------------------

# 4. Linear Regression Analysis

## 4.1 Introduction to the Problem

Predicting student grades is fundamental to early intervention strategies. Linear regression provides an interpretable framework for understanding how various factors—from prior academic preparation to socioeconomic conditions—combine to influence academic performance. While simple in concept, this approach reveals which factors matter most and by how much.

## 4.2 Methodology

Linear regression models the relationship between predictor variables and a continuous outcome through the equation:

$$Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + ... + \beta_nX_n + \epsilon$$

where $Y$ represents second semester grades, $X_i$ are our predictor variables, $\beta_i$ are coefficients indicating the strength and direction of each relationship, and $\epsilon$ captures random error.

We predict second semester grades based on 13 carefully selected features including demographic factors, previous academic performance, and economic indicators. This multi-factor approach recognizes that student success results from the interplay of personal, academic, and contextual variables.

## 4.3 Data Preparation

```{r linear-prep, results="hide"}
TARGET <- 'Curricular.units.2nd.sem..grade.'

FEATURES <- c(
    'Previous.qualification..grade.',
    'Admission.grade',
    'Age.at.enrollment',
    "Mother.s.qualification",
    "Father.s.qualification",
    'Scholarship.holder',
    'Curricular.units.1st.sem..credited.',
    'Curricular.units.1st.sem..enrolled.',
    'Curricular.units.1st.sem..approved.',
    'Curricular.units.1st.sem..grade.',
    'Unemployment.rate',
    'Inflation.rate',
    'GDP'
)

df_model <- df %>%
    select(all_of(c(FEATURES, TARGET))) %>%
    mutate(across(everything(), ~ as.numeric(as.character(.)))) %>%
    drop_na() %>%
    filter(get(TARGET) > 0)
```

After filtering missing values and excluding zero grades, the linear regression model uses **`r nrow(df_model)` students**. The target variable (**2nd semester grade**) has **mean** **`r round(mean(df_model[[TARGET]]), 2)`** and **standard deviation** **`r round(sd(df_model[[TARGET]]), 2)`**.

The dataset is filtered to **exclude students with zero grades**, as these typically represent withdrawals or incomplete coursework rather than actual performance. This ensures our model focuses on **predicting outcomes for actively enrolled students**.

## 4.4 Exploratory Analysis

```{r linear-eda, results="asis"}
correlations <- cor(df_model[FEATURES], df_model[[TARGET]])
correlations_sorted <- sort(correlations[,1], decreasing = TRUE)

top5_cor <- data.frame(
  Feature = names(head(correlations_sorted, 5)),
  Correlation = as.numeric(round(head(correlations_sorted, 5), 3))
)

knitr::kable(top5_cor, caption = "Top 5 correlations with 2nd semester grade")
```

### Feature Relationships

```{r linear-correlation, fig.height=8, fig.width=10}
correlation_matrix <- cor(df_model[c(FEATURES, TARGET)])

corrplot(correlation_matrix,
         method = "color",
         type = "upper",
         tl.col = "black",
         tl.srt = 45,
         tl.cex = 1.1,        # Increased from 0.7 for bigger text
         col = colorRampPalette(c("blue", "white", "red"))(200),
         addCoef.col = "black",
         number.cex = 0.9,    # Increased from 0.4 for readable numbers
         title = "Feature Correlation Heatmap",
         mar = c(0, 0, 2, 0))
```

**Key Observations:** Strong positive correlations are evident between **first semester performance metrics** and **second semester grades**, suggesting that prior academic success is highly predictive. This finding supports the importance of **early intervention programs** targeting first-semester students.

## 4.5 Model Development

```{r linear-split, results="hide"}
set.seed(42)

trainIndex <- createDataPartition(df_model[[TARGET]], p = 0.8, list = FALSE)

train_data <- df_model[trainIndex, ]
test_data <- df_model[-trainIndex, ]

X_train <- train_data %>% dplyr::select(all_of(FEATURES))
y_train <- train_data[[TARGET]]
X_test <- test_data %>% dplyr::select(all_of(FEATURES))
y_test <- test_data[[TARGET]]

train_n <- nrow(train_data)
test_n <- nrow(test_data)
```

We use an **80/20 train-test split**, resulting in **`r train_n`** training observations and **`r test_n`** test observations.

```{r linear-scale}
preProc <- preProcess(X_train, method = c("center", "scale"))
X_train_scaled <- predict(preProc, X_train)
X_test_scaled <- predict(preProc, X_test)
```

**Feature Scaling:** Standardization (**z-score normalization**) ensures all features contribute proportionally to the model regardless of their original scales.

```{r linear-train, results="asis", echo=FALSE}
train_data_scaled <- X_train_scaled
train_data_scaled[[TARGET]] <- y_train

formula_str <- paste(TARGET, "~", paste(FEATURES, collapse = " + "))
formula_obj <- as.formula(formula_str)

model <- lm(formula_obj, data = train_data_scaled)
coef_summary <- data.frame(
  Feature = names(coef(model)),
  Coefficient = coef(model)
)
coef_table <- coef_summary %>%
  dplyr::filter(Feature != "(Intercept)") %>%
  dplyr::mutate(Abs_Coefficient = abs(Coefficient)) %>%
  dplyr::arrange(desc(Abs_Coefficient)) %>%
  dplyr::select(Feature, Coefficient) %>%
  utils::head(3)

rownames(coef_table) <- NULL

if (requireNamespace("kableExtra", quietly = TRUE)) {
  knitr::kable(
    coef_table,
    caption = "Top 3 linear regression coefficients (by absolute value)",
    align = c("l", "r")
  ) %>%
    kableExtra::kable_styling(full_width = FALSE) %>%
    kableExtra::column_spec(1, bold = TRUE)
} else {
  knitr::kable(
    coef_table,
    caption = "Top 3 linear regression coefficients (by absolute value)",
    align = c("l", "r")
  )
}
```

## 4.6 Model Performance

```{r linear-predict, results="hide"}
y_train_pred <- predict(model, X_train_scaled)
y_test_pred <- predict(model, X_test_scaled)

train_r2 <- cor(y_train, y_train_pred)^2
train_rmse <- sqrt(mean((y_train - y_train_pred)^2))
train_mae <- mean(abs(y_train - y_train_pred))

test_r2 <- cor(y_test, y_test_pred)^2
test_rmse <- sqrt(mean((y_test - y_test_pred)^2))
test_mae <- mean(abs(y_test - y_test_pred))
```

**Performance Interpretation:** The R² score of `r round(test_r2, 4)` indicates that `r round(test_r2 * 100, 1)`% of variance in second semester grades can be explained by our selected features. The Mean Absolute Error (MAE) of `r round(test_mae, 2)` means our predictions are typically within `r round(test_mae, 2)` grade points of actual values.

## 4.7 Visual Assessment

```{r linear-plots, fig.cap="Actual vs. Predicted Grades"}
par(mfrow = c(1, 2))

plot(y_train, y_train_pred,
     xlab = "Actual Grade", ylab = "Predicted Grade",
     main = sprintf("Training Set (R² = %.4f)", train_r2),
     pch = 16, col = rgb(0, 0, 1, 0.5))
abline(a = 0, b = 1, col = "red", lwd = 2, lty = 2)
grid()

plot(y_test, y_test_pred,
     xlab = "Actual Grade", ylab = "Predicted Grade",
     main = sprintf("Test Set (R² = %.4f)", test_r2),
     pch = 16, col = rgb(0, 1, 0, 0.5))
abline(a = 0, b = 1, col = "red", lwd = 2, lty = 2)
grid()
```

### Residual Analysis

```{r linear-residuals, fig.cap="Residual diagnostics"}
residuals_test <- y_test - y_test_pred

par(mfrow = c(1, 2))

plot(y_test_pred, residuals_test,
     xlab = "Predicted Grade", ylab = "Residuals",
     main = "Residual Plot",
     pch = 16, col = rgb(0, 0, 1, 0.5))
abline(h = 0, col = "red", lwd = 2, lty = 2)
grid()

hist(residuals_test, breaks = 30,
     xlab = "Residuals", ylab = "Frequency",
     main = "Residual Distribution",
     col = rgb(0, 0, 1, 0.7), border = "black")
abline(v = 0, col = "red", lwd = 2, lty = 2)
grid()
```

## 4.8 Feature Importance

```{r linear-importance, fig.height=8, fig.cap="Feature importance"}
feature_importance <- data.frame(
    Feature = names(coef(model))[-1],
    Coefficient = coef(model)[-1]
) %>%
    mutate(Abs_Coefficient = abs(Coefficient)) %>%
    arrange(desc(Abs_Coefficient))

par(mar = c(5, 12, 4, 2))
colors <- ifelse(feature_importance$Coefficient > 0, "green", "red")
barplot(feature_importance$Coefficient,
        names.arg = feature_importance$Feature,
        horiz = TRUE,
        las = 1,
        col = adjustcolor(colors, alpha.f = 0.7),
        xlab = "Coefficient Value (Standardized)",
        main = "Feature Importance in Grade Prediction",
        xlim = c(min(feature_importance$Coefficient) * 1.2,
                 max(feature_importance$Coefficient) * 1.2))
abline(v = 0, col = "black", lwd = 1)
grid()
```

**Key Insights:** First semester grades have the strongest positive effect—students who perform well initially tend to continue succeeding. Course approvals and enrollments in the first semester significantly impact second semester outcomes. Economic indicators show measurable but smaller effects.

------------------------------------------------------------------------

# 5. Generalized Linear Model - Binomial

## 5.1 Introduction to Classification

Classification problems require predicting categorical outcomes rather than continuous values. Binomial logistic regression excels at binary classification by estimating probabilities between 0 and 1 through the logistic (sigmoid) function:

$$P(Y=1) = \frac{1}{1 + e^{-(\beta_0 + \beta_1X_1 + ... + \beta_nX_n)}}$$

This S-shaped curve ensures predictions stay within valid probability bounds while allowing us to model how predictor variables influence the odds of an outcome.

## 5.2 Binomial GLM: Graduation vs Dropout

**This is our most actionable model**—identifying students **at risk of dropping out** enables **targeted interventions**.

### Data Preparation and Model Training

```{r binomial2-all, results="asis", echo=FALSE}
# --- Data prep
df_success <- df %>%
  filter(Target %in% c('Graduate', 'Dropout')) %>%
  mutate(Success = ifelse(Target == 'Graduate', 1, 0))

cat(sprintf("Graduation dataset | Dropout=%d | Graduate=%d\n",
            sum(df_success$Success == 0),
            sum(df_success$Success == 1)))

features_success <- c(
  'Previous.qualification..grade.',
  'Admission.grade',
  'Age.at.enrollment',
  'Mother.s.qualification',
  'Father.s.qualification',
  'Scholarship.holder',
  'Gender',
  'Debtor',
  'Tuition.fees.up.to.date',
  'Curricular.units.1st.sem..approved.',
  'Curricular.units.1st.sem..grade.',
  'Unemployment.rate',
  'Inflation.rate',
  'GDP'
)

df_success_model <- df_success %>%
  select(all_of(c(features_success, 'Success'))) %>%
  mutate(across(everything(), as.numeric)) %>%
  na.omit()

cat(sprintf("Binomial (Graduation) dataset: %d students\n", nrow(df_success_model)))

# --- Train/test split + scaling
set.seed(42)
train_index_suc <- createDataPartition(df_success_model$Success, p = 0.8, list = FALSE)
train_suc <- df_success_model[train_index_suc, ]
test_suc <- df_success_model[-train_index_suc, ]

preproc_suc <- preProcess(train_suc[, features_success], method = c("center", "scale"))
train_suc_scaled <- predict(preproc_suc, train_suc)
test_suc_scaled <- predict(preproc_suc, test_suc)

# --- Model training
glm_suc <- glm(Success ~ .,
               data = train_suc_scaled,
               family = binomial(link = "logit"))

# --- Performance
pred_suc_prob <- predict(glm_suc, test_suc_scaled, type = "response")
pred_suc <- ifelse(pred_suc_prob > 0.5, 1, 0)

accuracy_suc <- mean(pred_suc == test_suc_scaled$Success)

cat(sprintf("Binomial GLM (Graduation) accuracy: %.4f (%.2f%%)\n", accuracy_suc, accuracy_suc * 100))

roc_suc <- roc(test_suc_scaled$Success, pred_suc_prob, quiet = TRUE)
auc_suc <- auc(roc_suc)

cat(sprintf("Binomial GLM (Graduation) AUC: %.3f\n", auc_suc))

# --- Risk/protective factors
coef_df_suc <- data.frame(
  Feature = features_success,
  Coefficient = coef(glm_suc)[-1],
  Odds_Ratio = exp(coef(glm_suc)[-1])
) %>%
  arrange(desc(abs(Coefficient)))

cat("\nTop 5 Predictors Interpretation:\n")
for (i in 1:min(5, nrow(coef_df_suc))) {
  row <- coef_df_suc[i, ]
  if (row$Odds_Ratio > 1) {
    pct_change <- (row$Odds_Ratio - 1) * 100
    cat(sprintf("✓ %s: %.1f%% increase in graduation odds per SD increase\n",
                row$Feature, pct_change))
  } else {
    pct_change <- (1 - row$Odds_Ratio) * 100
    cat(sprintf("✗ %s: %.1f%% decrease in graduation odds per SD increase\n",
                row$Feature, pct_change))
  }
}

# --- Top predictors table
odds_table_suc <- coef_df_suc %>%
  mutate(Abs_Coefficient = abs(Coefficient)) %>%
  arrange(desc(Abs_Coefficient)) %>%
  select(Feature, Coefficient, Odds_Ratio) %>%
  head(3)

knitr::kable(odds_table_suc, caption = "Top 3 graduation predictors (log-odds and odds ratio)")
```

We model graduation vs dropout in one step: build the binary target, split/scale, train a binomial GLM, and summarize discrimination (AUC) plus odds ratios for the most influential features. The model delivers **AUC = `r round(auc_suc, 3)`**, showing strong ability to predict graduates and dropouts. First-semester performance and financial factors are proven to be top factors.

**Actionable Insights:** **First semester performance** emerges as the dominant predictor. **Financial factors** (debtor status, tuition payment) also show strong effects, suggesting that financial aid could meaningfully impact retention.

------------------------------------------------------------------------

# 6. Generalized Linear Model - Poisson

## 6.1 Why Poisson Regression?

Course completion counts represent non-negative integers (0, 1, 2, 3...) that don't follow a normal distribution. Poisson regression solves this by:

1.  **Ensuring non-negative predictions** through a log link function
2.  **Modeling the count distribution** appropriately
3.  **Accounting for variance** that changes with the mean

The model structure is: $\log(\lambda) = \beta_0 + \beta_1X_1 + ... + \beta_nX_n$, where $\lambda$ represents the expected count.

## 6.2 Model 1: First Semester Course Approvals

Understanding what predicts first semester success is crucial for early intervention.

### Data Preparation

```{r poisson1-prep, results="hide"}
TARGET_1 <- 'Curricular.units.1st.sem..approved.'

FEATURES_1 <- c(
    'Previous.qualification..grade.',
    'Admission.grade',
    'Age.at.enrollment',
    "Mother.s.qualification",
    "Father.s.qualification",
    'Scholarship.holder',
    'Gender',
    'Debtor',
    'Tuition.fees.up.to.date',
    'Curricular.units.1st.sem..enrolled.',
    'Unemployment.rate',
    'GDP'
)

df_model1 <- df %>%
    select(all_of(c(FEATURES_1, TARGET_1))) %>%
    mutate(across(everything(), ~ as.numeric(as.character(.)))) %>%
    drop_na()

cat(sprintf("Poisson dataset (1st semester approvals): %d students\n", nrow(df_model1)))
set.seed(42)
X1 <- df_model1 %>% dplyr::select(all_of(FEATURES_1))
y1 <- df_model1[[TARGET_1]]

train_idx1 <- createDataPartition(y1, p = 0.8, list = FALSE)
X_train1 <- X1[train_idx1, ]
X_test1  <- X1[-train_idx1, ]
y_train1 <- y1[train_idx1]
y_test1  <- y1[-train_idx1]

train_data1 <- X_train1
train_data1[[TARGET_1]] <- y_train1

poisson_model1 <- glm(
  formula = as.formula(paste(TARGET_1, "~ .")),
  family = poisson(link = "log"),
  data = train_data1
)

# --- Performance
y_pred_train1 <- predict(poisson_model1, newdata = X_train1, type = "response")
y_pred_test1 <- predict(poisson_model1, newdata = X_test1, type = "response")

train_rmse1 <- sqrt(mean((y_train1 - y_pred_train1)^2))
train_mae1 <- mean(abs(y_train1 - y_pred_train1))

test_rmse1 <- sqrt(mean((y_test1 - y_pred_test1)^2))
test_mae1 <- mean(abs(y_test1 - y_pred_test1))
test_r2_1 <- cor(y_test1, y_pred_test1)^2

cat(sprintf("Poisson GLM performance | Train RMSE=%.4f MAE=%.4f | Test RMSE=%.4f MAE=%.4f R²=%.4f\n",
            train_rmse1, train_mae1, test_rmse1, test_mae1, test_r2_1))

# --- Rate ratios (top 3)
coef_summary1 <- summary(poisson_model1)$coefficients
coef_df1 <- data.frame(
    Feature = rownames(coef_summary1),
    Coefficient = coef_summary1[, "Estimate"],
    Rate_Ratio = exp(coef_summary1[, "Estimate"]),
    P_value = coef_summary1[, "Pr(>|z|)"]
) %>%
    mutate(Abs_Coef = abs(Coefficient)) %>%
    arrange(desc(Abs_Coef))

top_poisson_tbl <- coef_df1[, c("Feature", "Coefficient", "Rate_Ratio", "P_value")] %>%
  head(3)

knitr::kable(top_poisson_tbl, caption = "Top 3 Poisson GLM predictors (coefficient and rate ratio)")

# --- Coefficient bar plot (inline, compact)
coef_plot_df <- coef_df1 %>%
    filter(Feature != "(Intercept)") %>%
    head(10)

par(mar = c(5, 10, 3, 2), cex = 0.8)
colors <- ifelse(coef_plot_df$Coefficient > 0, "green", "red")
barplot(coef_plot_df$Coefficient,
        names.arg = coef_plot_df$Feature,
        horiz = TRUE,
        las = 1,
        col = adjustcolor(colors, alpha.f = 0.7),
        xlab = "Coefficient (Log Rate)",
        main = "Feature Effects on Course Approvals")
abline(v = 0, col = "black", lwd = 1)
grid()

# Store for plots below
assign("coef_df1", coef_df1, envir = .GlobalEnv)
assign("y_pred_train1", y_pred_train1, envir = .GlobalEnv)
assign("y_pred_test1", y_pred_test1, envir = .GlobalEnv)
assign("y_train1", y_train1, envir = .GlobalEnv)
assign("y_test1", y_test1, envir = .GlobalEnv)
```

**Interpreting Rate Ratios:** Rate ratios show how the expected course count changes per unit increase. For example, a rate ratio of **1.30** means **30% more courses** on average; **0.80** means **20% fewer**.

### Visualization of RMSE

```{r poisson1-plots, fig.cap="Model predictions for course approvals"}
par(mfrow = c(1, 2))

plot(y_train1, y_pred_train1,
     xlab = "Actual Courses", ylab = "Predicted Courses",
     main = sprintf("Training (RMSE = %.3f)",
                    sqrt(mean((y_train1 - y_pred_train1)^2))),
     pch = 16, col = rgb(0, 0, 1, 0.5))
abline(a = 0, b = 1, col = "red", lwd = 2, lty = 2)
grid()

plot(y_test1, y_pred_test1,
     xlab = "Actual Courses", ylab = "Predicted Courses",
     main = sprintf("Test (RMSE = %.3f)", test_rmse1),
     pch = 16, col = rgb(0, 1, 0, 0.5))
abline(a = 0, b = 1, col = "red", lwd = 2, lty = 2)
grid()
```

**RMSE reminder:** is the square-root of the average squared error; lower values mean predictions stay closer to actual course counts. Here the test RMSE shows the typical error distance (in courses) between predicted and actual approvals. The plots show a reasonable fit for common counts, but the model tends to underpredict for very high course approvals, so extreme tails are harder to capture.

------------------------------------------------------------------------

# 7. Generalized Additive Model - GAM

## 7.1 Beyond Linear Relationships

All our previous models assumed **linear relationships**. But reality is rarely this simple. **GAMs** extend **GLMs** by replacing straight lines with **smooth curves**, allowing the model to discover **non-linear patterns** while maintaining interpretability.

The model structure: $Y = \beta_0 + f_1(X_1) + f_2(X_2) + ... + f_n(X_n) + \epsilon$, where $f_i$ are smooth, flexible functions learned from data.

## 7.2 Model 1: Gaussian GAM for Grade Prediction

```{r gam1-all, results="hide"}
TARGET_GRADE <- 'Curricular.units.2nd.sem..grade.'

features_grade <- c(
  'Previous.qualification..grade.',
  'Admission.grade',
  'Age.at.enrollment',
  'Scholarship.holder',
  'Curricular.units.1st.sem..approved.',
  'Curricular.units.1st.sem..grade.',
  'Unemployment.rate',
  'GDP'
)

df_grade <- df %>%
  select(all_of(c(features_grade, TARGET_GRADE))) %>%
  mutate(across(everything(), as.numeric)) %>%
  na.omit() %>%
  filter(!!sym(TARGET_GRADE) > 0)

cat(sprintf("GAM (grade) dataset: %d students\n", nrow(df_grade)))

set.seed(42)
train_index_g <- createDataPartition(df_grade[[TARGET_GRADE]], p = 0.8, list = FALSE)
train_g <- df_grade[train_index_g, ]
test_g <- df_grade[-train_index_g, ]

gam_grade <- gam(
  Curricular.units.2nd.sem..grade. ~
    s(Previous.qualification..grade., bs = "cr", k = 10) +
    s(Admission.grade, bs = "cr", k = 10) +
    s(Age.at.enrollment, bs = "cr", k = 8) +
    factor(Scholarship.holder) +
    s(Curricular.units.1st.sem..approved., bs = "cr", k = 8) +
    s(Curricular.units.1st.sem..grade., bs = "cr", k = 10) +
    s(Unemployment.rate, bs = "cr", k = 8) +
    s(GDP, bs = "cr", k = 8),
  data = train_g,
  method = "REML"
)

cat("GAM summary:\n")
print(summary(gam_grade))

pred_test_g <- predict(gam_grade, test_g)

test_r2_g <- cor(test_g[[TARGET_GRADE]], pred_test_g)^2
test_rmse_g <- sqrt(mean((test_g[[TARGET_GRADE]] - pred_test_g)^2))
test_mae_g <- mean(abs(test_g[[TARGET_GRADE]] - pred_test_g))

cat(sprintf("GAM performance: R²=%.4f RMSE=%.4f MAE=%.4f\n",
            test_r2_g, test_rmse_g, test_mae_g))
```

### Comparison with Linear Model

```{r gam1-comparison, results="hide"}
lm_model <- lm(Curricular.units.2nd.sem..grade. ~ ., data = train_g)
lm_pred <- predict(lm_model, test_g)
lm_r2 <- cor(test_g[[TARGET_GRADE]], lm_pred)^2

cat(sprintf("GAM vs Linear | GAM R²=%.4f | Linear R²=%.4f | Improvement=%.2f pp\n",
            test_r2_g, lm_r2, (test_r2_g - lm_r2) * 100))
```

The GAM achieves an **R²** of `r round(test_r2_g, 4)` versus `r round(lm_r2, 4)` for the linear model—demonstrating that **non-linear relationships** exist and capturing them improves predictions.

### Visualizing Non-Linear Effects

```{r gam1-smooth, fig.height=8, fig.cap="Smooth functions reveal non-linear patterns"}
par(mfrow = c(2, 4), mar = c(4, 4, 3, 1))
plot(gam_grade, select = 1, shade = TRUE, col = "blue", shade.col = "lightblue",
     main = "Previous Qualification", cex.main = 1.2)
plot(gam_grade, select = 2, shade = TRUE, col = "blue", shade.col = "lightblue",
     main = "Admission Grade", cex.main = 1.2)
plot(gam_grade, select = 3, shade = TRUE, col = "blue", shade.col = "lightblue",
     main = "Age at Enrollment", cex.main = 1.2)
plot(gam_grade, select = 4, shade = TRUE, col = "blue", shade.col = "lightblue",
     main = "1st Sem Approved", cex.main = 1.2)
plot(gam_grade, select = 5, shade = TRUE, col = "blue", shade.col = "lightblue",
     main = "1st Sem Grade", cex.main = 1.2)
plot(gam_grade, select = 6, shade = TRUE, col = "blue", shade.col = "lightblue",
     main = "Unemployment Rate", cex.main = 1.2)
plot(gam_grade, select = 7, shade = TRUE, col = "blue", shade.col = "lightblue",
     main = "GDP", cex.main = 1.2)
```

**Key Non-Linear Patterns:** Age effects may show optimal ranges, course load might reveal sweet spots, and economic variables might display threshold effects that linear models cannot capture.

```{r gam1-predictions, fig.cap="GAM predictions maintain strong accuracy"}
par(mfrow = c(1, 2))

plot(test_g[[TARGET_GRADE]], pred_test_g,
     pch = 16, col = rgb(0.5, 0, 0.5, 0.5),
     xlab = "Actual Grade", ylab = "Predicted Grade",
     main = sprintf("GAM Predictions (R² = %.4f)", test_r2_g),
     cex.main = 1.3)
abline(0, 1, col = "red", lwd = 2, lty = 2)
grid()

residuals_g <- test_g[[TARGET_GRADE]] - pred_test_g
plot(pred_test_g, residuals_g,
     pch = 16, col = rgb(0.5, 0, 0.5, 0.5),
     xlab = "Predicted Grade", ylab = "Residuals",
     main = "Residual Plot",
     cex.main = 1.3)
abline(h = 0, col = "red", lwd = 2, lty = 2)
grid()
```

## 7.3 Model 2: Logistic GAM for Graduation Prediction

```{r gam2-prep}
df_success <- df %>%
  filter(Target %in% c('Graduate', 'Dropout')) %>%
  mutate(Success = ifelse(Target == 'Graduate', 1, 0))

features_success <- c(
  'Admission.grade',
  'Age.at.enrollment',
  'Scholarship.holder',
  'Tuition.fees.up.to.date',
  'Curricular.units.1st.sem..approved.',
  'Curricular.units.1st.sem..grade.',
  'Unemployment.rate',
  'GDP'
)

df_success_model <- df_success %>%
  dplyr::select(all_of(c(features_success, 'Success'))) %>%
  mutate(across(everything(), as.numeric)) %>%
  na.omit()

set.seed(42)
train_index_s <- createDataPartition(df_success_model$Success, p = 0.8, list = FALSE)
train_s <- df_success_model[train_index_s, ]
test_s <- df_success_model[-train_index_s, ]

gam_success <- gam(
  Success ~
    s(Admission.grade, bs = "cr", k = 10) +
    s(Age.at.enrollment, bs = "cr", k = 8) +
    factor(Scholarship.holder) +
    factor(Tuition.fees.up.to.date) +
    s(Curricular.units.1st.sem..approved., bs = "cr", k = 8) +
    s(Curricular.units.1st.sem..grade., bs = "cr", k = 10) +
    s(Unemployment.rate, bs = "cr", k = 8) +
    s(GDP, bs = "cr", k = 8),
  data = train_s,
  family = binomial(link = "logit"),
  method = "REML"
)
```

```{r gam2-eval, results="hide"}
pred_proba_s <- predict(gam_success, test_s, type = "response")
pred_s <- ifelse(pred_proba_s > 0.5, 1, 0)

accuracy_s <- mean(pred_s == test_s$Success)
roc_s <- roc(test_s$Success, pred_proba_s, quiet = TRUE)
auc_s <- auc(roc_s)

cat(sprintf("Accuracy: %.4f (%.2f%%)\n", accuracy_s, accuracy_s * 100))
cat(sprintf("AUC-ROC: %.4f\n", auc_s))
```

For the graduation GAM, the test set results are **accuracy = `r round(accuracy_s, 3)`** and **AUC = `r round(auc_s, 3)`**.

```{r gam2-comparison, results="hide"}
log_reg <- glm(Success ~ ., data = train_s, family = binomial)
log_pred_proba <- predict(log_reg, test_s, type = "response")
log_roc <- roc(test_s$Success, log_pred_proba, quiet = TRUE)
log_auc <- auc(log_roc)

cat("\nLogistic GAM vs Logistic GLM:\n")
cat(sprintf("GAM AUC:         %.4f\n", auc_s))
cat(sprintf("Logistic GLM:    %.4f\n", log_auc))
cat(sprintf("Improvement:     %.2f percentage points\n", (auc_s - log_auc) * 100))
```

The GAM's **AUC** of `r round(auc_s, 4)` versus `r round(log_auc, 4)` for standard logistic regression represents meaningful improvement for **intervention decisions**.

### Non-Linear Risk Patterns

```{r gam2-smooth, fig.height=8, fig.cap="Non-linear graduation probability patterns"}
par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))
plot(gam_success, select = 1, shade = TRUE, col = "darkgreen", shade.col = "lightgreen",
     main = "Admission Grade", cex.main = 1.2, ylab = "Log-Odds of Graduation")
plot(gam_success, select = 2, shade = TRUE, col = "darkgreen", shade.col = "lightgreen",
     main = "Age at Enrollment", cex.main = 1.2, ylab = "Log-Odds of Graduation")
plot(gam_success, select = 3, shade = TRUE, col = "darkgreen", shade.col = "lightgreen",
     main = "1st Sem Approved", cex.main = 1.2, ylab = "Log-Odds of Graduation")
plot(gam_success, select = 4, shade = TRUE, col = "darkgreen", shade.col = "lightgreen",
     main = "1st Sem Grade", cex.main = 1.2, ylab = "Log-Odds of Graduation")
plot(gam_success, select = 5, shade = TRUE, col = "darkgreen", shade.col = "lightgreen",
     main = "Unemployment", cex.main = 1.2, ylab = "Log-Odds of Graduation")
plot(gam_success, select = 6, shade = TRUE, col = "darkgreen", shade.col = "lightgreen",
     main = "GDP", cex.main = 1.2, ylab = "Log-Odds of Graduation")
```

```{r gam2-roc, fig.cap="Logistic GAM achieves strong discrimination"}
plot(roc_s, col = "darkgreen", lwd = 2,
     main = sprintf("ROC Curve - Logistic GAM (AUC = %.3f)", auc_s),
     cex.main = 1.3)
abline(a = 0, b = 1, lty = 2, lwd = 2, col = "black")
legend("bottomright", legend = c("GAM", "Random"),
       col = c("darkgreen", "black"), lwd = 2, lty = c(1, 2))
grid()
```



------------------------------------------------------------------------

# 8. Neural Network Analysis

We use neural networks as an **early warning system** to predict whether a student is at risk of **dropping out** (**Dropout vs NotDropout**) based on early academic performance and financial indicators. Here, **NotDropout** means the student is either **still enrolled** or has **graduated**. The goal is not only accuracy, but also **actionability**: the model should highlight which signals (grades, approved courses, tuition status) are most associated with dropout risk.

## 8.1 Objective

We want our neural network to play the role of an early-warning system: predict **Dropout vs NotDropout** by learning the messy, non-linear patterns hidden in the data.

## 8.2 Feature Selection

We focus on three feature blocks from the preprocessed dataset:

1. **Early academic performance:** 1st- and 2nd-semester grades plus approved credits showed the highest F-values in our ANOVA screening during preprocessing, making them the strongest predictors of the target variable.
2. **Financial pressure/support:** Tuition fees up to date, debtor status, and scholarship holder indicate whether financial stress may affect the student.
3. **Entry profile:** Admission grade and age at enrollment provide baseline information about the student's starting point.

## 8.3 Data Preprocessing for Neural Networks

```{r nn-load-data, include=FALSE}
# Create binary target: Dropout vs NotDropout (Enrolled + Graduate)
df_nn <- df
df_nn$Target <- ifelse(df_nn$Target == "Dropout", "Dropout", "NotDropout")
df_nn$Target <- factor(df_nn$Target, levels = c("Dropout", "NotDropout"))

# Selected features for neural network
nn_features <- c(
  "Curricular.units.1st.sem..approved.",
  "Curricular.units.1st.sem..grade.",
  "Curricular.units.2nd.sem..approved.",
  "Curricular.units.2nd.sem..grade.",
  "Tuition.fees.up.to.date",
  "Scholarship.holder",
  "Age.at.enrollment",
  "Admission.grade",
  "Debtor"
)

# Create modeling dataset
nn_model_data <- df_nn %>%
  select(all_of(nn_features), Target) %>%
  na.omit()

# Scaling (important for neural networks)
numeric_cols <- sapply(nn_model_data, is.numeric)
nn_model_data[, numeric_cols] <- scale(nn_model_data[, numeric_cols])
```

```{r nn-data-overview, fig.height=4, results="hide"}
target_counts <- nn_model_data %>%
  count(Target) %>%
  mutate(Percent = n / sum(n) * 100)

ggplot(target_counts, aes(x = Target, y = n, fill = Target)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = paste0(n, " (", round(Percent, 1), "%)")), vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Dropout" = "#E74C3C", "NotDropout" = "#27AE60")) +
  labs(
    title = "Target distribution in the neural network dataset",
    x = NULL,
    y = "Number of students"
  ) +
  theme(legend.position = "none")
```

This modeling subset contains **`r nrow(nn_model_data)`** students after removing missing values. The chart above shows the class balance.

## 8.4 Train/Test Split

```{r nn-train-test-split, results="hide"}
set.seed(123)
train_idx <- createDataPartition(nn_model_data$Target, p = 0.8, list = FALSE)
train_nn <- nn_model_data[train_idx, ]
test_nn <- nn_model_data[-train_idx, ]
```

We use an **80/20 split**: training = **`r nrow(train_nn)`**, test = **`r nrow(test_nn)`**.

## 8.5 Model Training

We train two neural networks to compare Lab 1 and Lab 2 approaches.

### Model A: Single Hidden Layer (Lab 1 - nnet)

Following `ANN_Lab1_GermanCredit_v1.Rmd`, we use `nnet` with one hidden layer and weight decay.

```{r nn-train-model-A, results="hide", cache=TRUE, cacherebuild=FALSE}
# Model A: nnet with single hidden layer (Lab 1 style)
model_A <- nnet(
  Target ~ .,
  data = train_nn,
  size = 10,        # 10 hidden neurons in one layer
  decay = 0.01,     # Weight decay: small penalty on large weights to prevent overfitting
  maxit = 500,      # Maximum training iterations
  trace = FALSE     # Suppress training progress output
)
```

### Model B: Three Hidden Layers (Lab 2 - neuralnet)

Following `ANN_Lab2_Iris_v1.Rmd`, we use `neuralnet` with a deeper architecture using **three hidden layers**. This requires numeric inputs and numeric output columns, so we build 0/1 columns for our binary target.

In this run, Model B uses a **2-5-2** hidden-layer setup (2 neurons → 5 neurons → 2 neurons). We also set `threshold = 0.8` mainly to keep runtime manageable while we debug and iterate. A **larger** threshold usually stops earlier (faster), but it can also stop before the model has learned enough.

```{r nn-train-model-B, results="hide", cache=TRUE, cache.rebuild=FALSE}
# One-hot encode predictors for neuralnet (Lab 2 approach)
dummies <- dummyVars(Target ~ ., data = train_nn)
train_matrix <- as.data.frame(predict(dummies, newdata = train_nn))
test_matrix <- as.data.frame(predict(dummies, newdata = test_nn))

# Create dummy columns for binary target (neuralnet needs numeric output)
train_matrix$Target_Dropout <- as.numeric(train_nn$Target == "Dropout")
train_matrix$Target_NotDropout <- as.numeric(train_nn$Target == "NotDropout")

# Build formula: output columns ~ input columns
input_names <- names(train_matrix)[!grepl("Target", names(train_matrix))]
output_names <- c("Target_Dropout", "Target_NotDropout")
formula_B <- as.formula(paste(
  paste(output_names, collapse = " + "), "~",
  paste(input_names, collapse = " + ")
))

# Model B: neuralnet with three hidden layers (Lab 2 style)
model_B <- neuralnet(
  formula_B,
  data = train_matrix,
  hidden = c(2,5,2),   
  linear.output = FALSE,
  threshold = 0.8, 
  stepmax = 3e5      # enough iterations without being too slow
)
```

## 8.6 Model Evaluation

We generate predictions and confusion matrices for both models.

```{r nn-evaluate-both-models, results="hide"}
# --- Model A predictions (nnet) ---
pred_A_class <- predict(model_A, newdata = test_nn, type = "class")
pred_A_class <- factor(pred_A_class, levels = levels(test_nn$Target))
conf_matrix_A <- confusionMatrix(pred_A_class, test_nn$Target)

# --- Model B predictions (neuralnet) ---
# compute() runs the network on new data and returns raw output values
# Must convert to matrix (neuralnet requires numeric matrix, not data.frame)
pred_B_raw <- compute(model_B, as.matrix(test_matrix[, input_names]))$net.result
# Pick the class with highest output value for each student
pred_B_class <- factor(
  c("Dropout", "NotDropout")[apply(pred_B_raw, 1, which.max)],
  levels = c("Dropout", "NotDropout")
)
conf_matrix_B <- confusionMatrix(pred_B_class, test_nn$Target)
```

## 8.7 Performance Metrics Summary

We report common classification metrics for both models. In the binary setting, we treat **Dropout** as the "positive" class (this matches the business goal of flagging students who may need support).

```{r nn-performance-summary, results="hide"}
nn_metrics_df <- tibble(
  Model = c("Model A: nnet (1 hidden layer)", "Model B: neuralnet (3 hidden layers)"),
  Accuracy = c(
    as.numeric(conf_matrix_A$overall["Accuracy"]),
    as.numeric(conf_matrix_B$overall["Accuracy"])
  ),
  Precision = c(
    as.numeric(conf_matrix_A$byClass["Pos Pred Value"]),
    as.numeric(conf_matrix_B$byClass["Pos Pred Value"])
  ),
  Recall = c(
    as.numeric(conf_matrix_A$byClass["Sensitivity"]),
    as.numeric(conf_matrix_B$byClass["Sensitivity"])
  )
)
```

```{r nn-performance-dashboard, fig.height=4, fig.width=12}
nn_metrics_long <- nn_metrics_df %>%
  pivot_longer(cols = c(Accuracy, Precision, Recall), names_to = "Metric", values_to = "Value") %>%
  mutate(
    Metric = recode(
      Metric,
      Accuracy = "Accuracy",
      Precision = "Precision (Dropout)",
      Recall = "Recall (Dropout)"
    )
  )

ggplot(nn_metrics_long, aes(x = Metric, y = Value, fill = Model)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65, alpha = 0.9) +
  geom_text(
    aes(label = sprintf("%.3f", Value)),
    position = position_dodge(width = 0.75),
    vjust = -0.4,
    size = 4
  ) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("#5DADE2", "#48C9B0")) +
  labs(
    title = "Neural network performance on the test set",
    x = NULL,
    y = "Score",
    fill = "Model"
  ) +
  theme(legend.position = "top")
```

Model A (1 layer) still delivers the strongest result on the test set. The deeper Model B (hidden layers: 2-5-2) does not automatically outperform the simpler Model A. This mirrors the lecture message that more layers are not automatically better for structured tabular data.

## 8.8 Confusion Matrix Comparison

Side-by-side heatmaps comparing both models. X-axis (top) = Predicted, Y-axis = True labels.

```{r nn-confusion-matrix-comparison, fig.width=12, fig.height=5}
# Helper function to create confusion matrix plot
nn_plot_confusion_matrix <- function(conf_matrix, title) {
  cm_df <- as.data.frame(conf_matrix$table)
  names(cm_df) <- c("Predicted", "True", "Count")
  
  ggplot(cm_df, aes(x = Predicted, y = True, fill = Count)) +
    geom_tile(color = "white", linewidth = 1) +
    geom_text(aes(label = Count), size = 5, fontface = "bold", color = "white") +
    scale_fill_gradient(low = "#3498DB", high = "#E74C3C", name = "Count") +
    scale_x_discrete(position = "top") +
    scale_y_discrete(limits = rev(c("Dropout", "NotDropout"))) +
    labs(title = title, x = "Predicted Class", y = "True Class") +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
      axis.text = element_text(size = 10),
      legend.position = "right"
    )
}

# Create plots for both models
plot_A <- nn_plot_confusion_matrix(conf_matrix_A, "Model A: nnet (1 hidden layer)")
plot_B <- nn_plot_confusion_matrix(conf_matrix_B, "Model B: neuralnet (3 hidden layers)")

# Display side by side
grid.arrange(plot_A, plot_B, ncol = 2)
```

## 8.9 ROC Curve Comparison

We compare ROC curves for **Dropout** (Dropout vs NotDropout) across both models using `ROCR`.

```{r nn-roc-comparison, results="hide"}
# --- Model A ROC (nnet) ---
prob_A <- predict(model_A, newdata = test_nn, type = "raw")

# Safety: nnet with binary target outputs P(Class 2), which is NotDropout.
# We need P(Dropout).
if (is.null(dim(prob_A)) || ncol(prob_A) == 1) {
  prob_A_drop <- 1 - as.numeric(prob_A)
} else if ("Dropout" %in% colnames(prob_A)) {
  prob_A_drop <- prob_A[, "Dropout"]
} else {
  prob_A_drop <- prob_A[, 1]
}

pred_A_drop <- ROCR::prediction(prob_A_drop, test_nn$Target == "Dropout")
perf_A_drop <- ROCR::performance(pred_A_drop, "tpr", "fpr")
auc_A <- round(ROCR::performance(pred_A_drop, "auc")@y.values[[1]], 3)

# --- Model B ROC (neuralnet) ---
if (is.null(dim(pred_B_raw))) {
  prob_B_drop <- as.numeric(pred_B_raw)
} else {
  prob_B_drop <- pred_B_raw[, 1]
}
pred_B_drop <- ROCR::prediction(prob_B_drop, test_nn$Target == "Dropout")
perf_B_drop <- ROCR::performance(pred_B_drop, "tpr", "fpr")
auc_B <- round(ROCR::performance(pred_B_drop, "auc")@y.values[[1]], 3)
```

```{r nn-roc-comparison-plot, fig.height=4.5, fig.width=8}
roc_A_df <- data.frame(
  FPR = perf_A_drop@x.values[[1]],
  TPR = perf_A_drop@y.values[[1]],
  Model = sprintf("Model A (AUC=%.3f)", auc_A)
)
roc_B_df <- data.frame(
  FPR = perf_B_drop@x.values[[1]],
  TPR = perf_B_drop@y.values[[1]],
  Model = sprintf("Model B (AUC=%.3f)", auc_B)
)
roc_df <- bind_rows(roc_A_df, roc_B_df)

ggplot(roc_df, aes(x = FPR, y = TPR, color = Model)) +
  geom_line(linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray50") +
  scale_color_manual(values = c("#2ecc71", "#3498db")) +
  labs(
    title = "ROC curve (Dropout vs NotDropout)",
    x = "False Positive Rate",
    y = "True Positive Rate",
    color = NULL
  ) +
  theme(legend.position = "bottom")
```

**AUC interpretation:** higher AUC means better separation of Dropout vs NotDropout. Here, Model A has AUC **`r auc_A`** and Model B has AUC **`r auc_B`**. But **the difference is minimal**.

## 8.10 Feature Importance (Permutation)

Permutation importance (same spirit as Lab 1's "what-if" checks) measures how much each *model's accuracy* drops when we randomly shuffle one predictor at a time. Bigger drops mean the learner *relied heavily on that feature*. It shows that **second semester grades are a very important factor**. It is interesting that at HSLU, we also feel like the **second semester is the hardest** and requires a lot of **effort and attention**.

```{r nn-permutation-importance, message=FALSE, results="hide"}
# Helper to compute accuracy drop for a single feature and model
permute_feature <- function(model_type, var_name) {
  permuted <- test_nn
  permuted[[var_name]] <- sample(permuted[[var_name]])
  
  if (model_type == "Model A: nnet (1 layer)") {
    baseline <- mean(predict(model_A, newdata = test_nn, type = "class") == test_nn$Target)
    perm_acc <- mean(predict(model_A, newdata = permuted, type = "class") == test_nn$Target)
  } else {
    permuted_matrix <- as.data.frame(predict(dummies, newdata = permuted))
    permuted_matrix[] <- lapply(permuted_matrix, as.numeric)
    class_map <- c("Dropout", "NotDropout")
    baseline <- mean(pred_B_class == test_nn$Target)
    perm_raw <- compute(model_B, as.matrix(permuted_matrix[, input_names]))$net.result
    perm_class <- factor(class_map[apply(perm_raw, 1, which.max)], levels = class_map)
    perm_acc <- mean(perm_class == test_nn$Target)
  }
  
  tibble(
    Model = model_type,
    Feature = var_name,
    AccuracyDrop = baseline - perm_acc
  )
}

perm_results <- map_dfr(
  nn_features,
  ~bind_rows(
    permute_feature("Model A: nnet (1 layer)", .x),
    permute_feature("Model B: neuralnet (3 layers)", .x)
  )
)

# Friendlier labels for the plot
feature_labels <- c(
  "Curricular.units.1st.sem..approved." = "1st Sem Credits Approved",
  "Curricular.units.1st.sem..grade." = "1st Sem Grade",
  "Curricular.units.2nd.sem..approved." = "2nd Sem Credits Approved",
  "Curricular.units.2nd.sem..grade." = "2nd Sem Grade",
  "Tuition.fees.up.to.date" = "Tuition Fees Up to Date",
  "Scholarship.holder" = "Scholarship Holder",
  "Age.at.enrollment" = "Age at Enrollment",
  "Admission.grade" = "Admission Grade",
  "Debtor" = "Debtor Status"
)

perm_results <- perm_results %>%
  mutate(Feature = recode(Feature, !!!feature_labels)) %>%
  group_by(Model) %>%
  mutate(Feature = fct_reorder(Feature, AccuracyDrop)) %>%
  ungroup()

ggplot(perm_results, aes(x = AccuracyDrop, y = Feature, fill = Model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Model, scales = "free_y") +
  labs(
    title = "Permutation Importance (Accuracy Drop when Shuffled)",
    x = "Accuracy Drop",
    y = "Feature"
  ) +
  scale_fill_manual(values = c("#5DADE2", "#48C9B0"))
```

```{r nn-permutation-importance-table, results="asis"}
perm_top <- perm_results %>%
  group_by(Model) %>%
  slice_max(order_by = AccuracyDrop, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(AccuracyDropPctPoints = round(AccuracyDrop * 100, 2)) %>%
  select(Model, Feature, AccuracyDropPctPoints)

knitr::kable(
  perm_top,
  caption = "Top 5 permutation importances (how many accuracy percentage points we lose when we shuffle the feature)"
)
```

## 8.11 Business Use: Early Warning and Intervention

This neural network can be used as an **early warning system**:

1. At the end of the 1st/2nd semester, we collect the same inputs (grades, approved units, tuition status, scholarship, debtor status).
2. The model outputs a **dropout risk score** for each student (the model's predicted output/probability for the **Dropout** class).
3. The university can then **target limited support resources** to students with the highest predicted risk.

Examples of interventions (linked to our most important features):

- **Academic performance signals (grades, approved units):** provide tutoring, mentoring, and early academic advising.
- **Financial stress signals (tuition up to date, debtor):** proactive outreach about payment plans and financial aid options.
- **Support signals (scholarship holder):** consider expanding support programs for similar students who are not currently covered.

Important caution:

- The model helps us **prioritize** students for support, but it should not be the only decision tool.
- The safest process is to combine model predictions with human review (advisors) before taking action.

## 8.12 Neural Network Architecture Visualization

`NeuralNetTools::plotnet()` turns the fitted networks into diagrams that show layers, neurons, and connection strengths (thicker lines = larger absolute weights).

```{r nn-visualization, fig.width=12, fig.height=6}
short_labels <- c("1st\nCredits", "1st\nGrade", "2nd\nCredits", "2nd\nGrade",
                  "Tuition\nOK", "Scholar", "Age", "Adm\nGrade", "Debtor")
output_labels_B <- c("Dropout", "NotDropout")

par(mfrow = c(1, 2), mar = c(2, 2, 2, 2))
NeuralNetTools::plotnet(
  model_A,
  alpha = 0.7,
  circle_col = list("steelblue", "skyblue"),
  x_names = short_labels,
  max_sp = TRUE,
  cex_val = 0.9,
  line_lwd = 0.8
)
NeuralNetTools::plotnet(
  model_B,
  alpha = 0.7,
  circle_col = list("darkseagreen", "palegreen"),
  x_names = short_labels,
  y_names = output_labels_B,
  max_sp = TRUE,
  cex_val = 0.9,
  line_lwd = 0.8
)
par(mfrow = c(1, 1))
```

## 8.13 Findings & Interpretation

**Model A (nnet):** 9 inputs - 10 hidden neurons - 1 output (P(NotDropout)). **Simple, fast**, handles factors directly and reaches strong test performance.

**Model B (neuralnet):** 9 inputs - 2 - 5 - 2 neurons - 2 outputs. Shows Lab 2's deeper setup but **does not automatically outperform the simpler Model A**.

**Strengths observed:** Neural networks capture non-linear interaction between grades, tuition status, and scholarships.

**Limitations observed:**

- Deeper networks remain sensitive to hyperparameters and can overfit small samples.
- Interpretability is limited compared to GLM/GAM coefficients.
- Because we combine **Enrolled + Graduate** into **NotDropout**, this target cannot distinguish "still studying" from "already graduated". This is acceptable for an early-warning dropout use case, but it matters if the business question is specifically "who will graduate".

## 8.14 Neural Network Conclusions

1. Both neural networks predict **dropout risk** reasonably well, but the single hidden layer (Model A) provides the best balance of accuracy and stability on this dataset.
2. Adding more layers (Model B: 2-5-2) does not automatically outperform the simpler architecture; this reinforces the lab insight that "deeper" is not automatically "better" for structured tabular data.
3. Standard preprocessing steps (scaling numeric features, stratified train/test split) remain essential for consistent results.

------------------------------------------------------------------------

# 9. Support Vector Machine Analysis

```{r svm-load-data, echo=TRUE}
# Load the dataset from the project data folder
student_data <- read_csv("data/preprocessed_data.csv")

# Create binary target: Dropout vs Not Dropout (Graduate + Enrolled)
# This aligns with the Early Warning System goal and is consistent with our Neural Network
student_data <- student_data %>%
  mutate(Target = ifelse(Target == "Dropout", "Dropout", "Not Dropout"))

# Convert Target to factor
student_data$Target <- factor(student_data$Target, levels = c("Dropout", "Not Dropout"))
```

We will now test different **SVM kernels** to understand which is the best classifier for our student performance prediction. SVMs operate by identifying where the **margin** between different categories of data is the greatest.
To classify our dataset, the model plots a separating line between the two outcome classes using a number of predictors - in order to keep it clear, we want to create a two-dimensional plane (two predictors).

By applying different kernels—linear, radial, and polynomial— we assess how the shape of the decision boundary affects prediction accuracy and determine which kernel provides the most reliable classification for our dataset.

## 9.1 Feature Selection

For us to choose the best predictors, we will compute **F-statistic ANOVA** for every numeric feature in the dataset. The F-score tells us _how much_ the **mean** of each feature **differs** between the **two target groups** (_Dropout_ vs _Not Dropout_). Higher scores mean bigger separation between classes, making the feature more useful for classification. 

To visualize the SVM decision boundary, we need to limit our analysis to **two features** (creating a 2D plot with X and Y axes). We selected **second semester marks** and **courses approved** as they have the highest F-values, indicating they best separate Dropout from Not Dropout students.
```{r svm-fscore-top5, echo=TRUE}
numeric_cols <- sapply(student_data, is.numeric)
features <- names(student_data)[numeric_cols]
# Remove Target if it's numeric
features <- features[features != "Target"]

results <- data.frame(
  Feature = character(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (feat in features) {
  formula <- as.formula(paste(feat, "~ Target"))
  anova_result <- aov(formula, data = student_data)
  summary_anova <- summary(anova_result)
  
  results <- rbind(
    results,
    data.frame(
      Feature = feat,
      F_value = round(summary_anova[[1]]$`F value`[1], 2),
      p_value = sprintf("%.3f", summary_anova[[1]]$`Pr(>F)`[1])
    )
  )
}

results <- results[order(-results$F_value), ]
rownames(results) <- NULL
knitr::kable(head(results, 5), caption = "Top 5 Features by F-statistic")
```

## 9.2 Data Preparation & Train/Test Split

For consistency with our Neural Network model and the Early Warning System goal, we use a **binary classification**: *Dropout* vs *Not Dropout* (which includes both Graduate and Enrolled students). This approach makes sense because at prediction time, all students are enrolled — we want to predict if they will drop out.
```{r svm-prepare-marks, echo=TRUE, results='hide'}
# Get the top 2 features (excluding Target)
top_feature_1 <- as.character(results$Feature[1])
top_feature_2 <- as.character(results$Feature[2])

# Create marks dataframe with selected features
marks <- data.frame(
  x.1 = student_data[[top_feature_1]],
  x.2 = student_data[[top_feature_2]],
  y = as.factor(student_data$Target)
)

# Remove rows with NA values if any exist
marks <- na.omit(marks)
```
```{r svm-prepare-marks-display, echo=TRUE}
# Create summary table
data_summary <- data.frame(
  Metric = c("Total Observations", "Target Classes"),
  Value = c(nrow(marks), paste(levels(marks$y), collapse=", "))
)

knitr::kable(data_summary, align = 'c', caption = "Dataset Summary")
```

We split the data 80/20 for training and testing:
```{r svm-train-test-split-display, echo=TRUE}
# svm-train-test-split
set.seed(123)
trainIndex <- createDataPartition(marks$y, p = 0.8, list = FALSE)
train_data <- marks[trainIndex, ]
test_data <- marks[-trainIndex, ]
# Create split summary table
split_summary <- data.frame(
  Dataset = c("Training", "Testing"),
  Observations = c(nrow(train_data), nrow(test_data)),
  Percentage = c(
    paste0(round(nrow(train_data)/nrow(marks)*100, 1), "%"),
    paste0(round(nrow(test_data)/nrow(marks)*100, 1), "%")
  )
)

knitr::kable(split_summary, align = 'c', caption = "Train/Test Split")
```

## 9.3 Model Training & Tuning

We train and optimize **all three SVM kernels** using `tune()` to find the best hyperparameters for each. This ensures a fair comparison rather than using arbitrary parameter values.

**Linear Kernel**: Creates a straight decision boundary to separate classes. This kernel is ideal when the data is linearly separable.  
**Radial Basis Function (RBF) Kernel**: Generates curved, non-linear decision boundaries by measuring the distance between data points.  
**Polynomial Kernel**: Produces curved boundaries with a specific degree of curve. 
```{r svm-train-models, echo=TRUE}
set.seed(123)

# 1. Linear Kernel (Tune cost only)
tune_linear <- tune(svm, y ~ x.1 + x.2, data = train_data,
                    kernel = "linear",
                    ranges = list(cost = c(0.1, 1, 10, 100)))
svm_linear <- tune_linear$best.model

# 2. Radial Kernel (Tune cost + gamma)
tune_radial <- tune(svm, y ~ x.1 + x.2, data = train_data, 
                    kernel = "radial",
                    ranges = list(cost = c(0.1, 1, 10, 100),
                                 gamma = c(0.1, 0.5, 1, 2)))
svm_radial <- tune_radial$best.model

# 3. Polynomial Kernel (Tune cost + degree)
tune_poly <- tune(svm, y ~ x.1 + x.2, data = train_data,
                  kernel = "polynomial",
                  ranges = list(cost = c(0.1, 1, 10),
                               degree = c(2, 3, 4)))
svm_poly <- tune_poly$best.model

```

**Optimal Hyperparameters:**

*   **Linear:** Cost = `r tune_linear$best.parameters$cost`. A lower cost suggests the data is not perfectly separable, allowing some misclassifications to find a wider margin (Soft Margin).
*   **Radial:** Cost = `r tune_radial$best.parameters$cost`, Gamma = `r tune_radial$best.parameters$gamma`. The gamma value controls the "reach" of a single training example; a moderate value implies a balance between capturing complex boundaries and overfitting.
*   **Polynomial:** Cost = `r tune_poly$best.parameters$cost`, Degree = `r tune_poly$best.parameters$degree`. The degree indicates the complexity of the curved boundary needed.

## 9.4 Decision Boundary Comparison

The plots below compare the **"Rigid" Linear Boundary** against the **"Flexible" Tuned Radial Boundary**. The decision boundary represents the "Safe Zone" — students above the line are predicted to *Not Dropout*, while those below are flagged as *Dropout* risk.

```{r svm-boundary-comparison, fig.width=10, fig.height=5, fig.align='center', echo=FALSE}
# Setup grid for plotting
x1_range <- range(train_data$x.1)
x2_range <- range(train_data$x.2)
x1_seq <- seq(x1_range[1], x1_range[2], length.out = 200)
x2_seq <- seq(x2_range[1], x2_range[2], length.out = 200)
grid <- expand.grid(x.1 = x1_seq, x.2 = x2_seq)

# Colors: Yellow (Dropout) vs Blue (Not Dropout)
cols <- c("#FFF9C4", "#B3E5FC")
pt_cols <- c("#FFA000", "#1976D2")

# Function to plot boundary
plot_svm_boundary <- function(model, title_text) {
  # Background Grid (Prediction)
  grid$pred <- predict(model, grid)
  plot(grid$x.1, grid$x.2, col = cols[as.numeric(grid$pred)], pch = 15, cex = 0.3,
       xlab = gsub("\\.", " ", top_feature_2),
       ylab = gsub("\\.", " ", top_feature_1),
       main = title_text,
       cex.lab = 0.9, cex.axis = 0.8)
  
  # Overlay TEST Data points (Actual)
  # Jitter added to separate overlapping integer scores
  points(jitter(test_data$x.1, amount = 0.3), jitter(test_data$x.2, amount = 0.3), 
         col = pt_cols[as.numeric(test_data$y)],
         pch = 19, cex = 0.6)
}

# Dual Plot Layout
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot_svm_boundary(svm_linear, "Linear Kernel (Test Data)")
legend("topright", legend = levels(test_data$y), col = pt_cols, pch = 19, cex = 0.7, bty = "n")

plot_svm_boundary(svm_radial, "Tuned Radial Kernel (Test Data)")
legend("topright", legend = levels(test_data$y), col = pt_cols, pch = 19, cex = 0.7, bty = "n")

par(mfrow = c(1, 1))
```


The **Linear kernel** creates a simple straight line cut-off. The **Radial kernel** is more sophisticated: notice how it creates a second "risk zone" (yellow blob) for students with many approved units but very low grades. The Linear model misses these at-risk students, classifying them as safe.

## 9.5 Visual Comparison: Confusion Matrices

Here we visually compare where the models make mistakes. The **Linear Model** serves as our baseline.

```{r svm-matrix-comparison, fig.width=10, fig.height=4, fig.align='center', echo=FALSE}
# Helper to plot matrix
plot_cm_tile <- function(pred_vals, title) {
  cm <- caret::confusionMatrix(pred_vals, test_data$y)
  cm_df <- as.data.frame(cm$table)
  colnames(cm_df) <- c("Predicted", "Actual", "Count")
  
  ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Count)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Count), color = "black", size = 5) +
    scale_fill_gradient(low = "#E3F2FD", high = "#2196F3") +
    labs(title = title, x = "Predicted", y = "Actual") +
    theme_minimal() +
    theme(legend.position = "none", plot.title = element_text(hjust = 0.5, face="bold"))
}

# Generate plots
pred_lin <- predict(svm_linear, test_data)
pred_rad <- predict(svm_radial, test_data)

p1 <- plot_cm_tile(pred_lin, "Linear Confusion Matrix")
p2 <- plot_cm_tile(pred_rad, "Radial Confusion Matrix")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**Interpretation:** The confusion matrices show where each model makes errors. For an Early Warning System, **false negatives** (predicting "Not Dropout" for students who actually dropout) are more costly than false positives.

Comparing the two models:

- **Linear kernel:** Catches 157 dropouts, but **misses 127** (false negatives)
- **Radial kernel:** Catches 195 dropouts, but **misses only 89** (false negatives)

The Radial kernel reduces missed dropout cases by **30%** (from 127 to 89). The trade-off is 26 extra false alarms (57 vs 31), which is acceptable since extra support for non-at-risk students causes no harm.

## 9.6 Performance Summary

All three kernels were optimized using grid search to find the best hyperparameters. The table below compares their final performance on the test set:

```{r svm-comparison, echo=TRUE}
# Calculate predictions for all models
pred_lin <- predict(svm_linear, test_data)
pred_rad <- predict(svm_radial, test_data)
pred_poly <- predict(svm_poly, test_data)

cm_linear <- confusionMatrix(pred_lin, test_data$y)
cm_radial <- confusionMatrix(pred_rad, test_data$y)
cm_poly   <- confusionMatrix(pred_poly, test_data$y)

comparison_results <- data.frame(
  Model = c("Linear", "Radial", "Polynomial"),
  Accuracy = c(cm_linear$overall['Accuracy'],
               cm_radial$overall['Accuracy'],
               cm_poly$overall['Accuracy']),
  Sensitivity = c(cm_linear$byClass['Sensitivity'],
                  cm_radial$byClass['Sensitivity'],
                  cm_poly$byClass['Sensitivity']),
  Kappa = c(cm_linear$overall['Kappa'],
            cm_radial$overall['Kappa'],
            cm_poly$overall['Kappa'])
)

# Store best model info using numeric values
svm_best_idx <- which.max(comparison_results$Accuracy)
svm_best_model <- comparison_results$Model[svm_best_idx]
svm_best_accuracy <- comparison_results$Accuracy[svm_best_idx]

# Create display table with formatting
display_table <- comparison_results
display_table[,2:3] <- lapply(display_table[,2:3], function(x) sprintf("%.1f%%", x*100))
display_table$Kappa <- round(display_table$Kappa, 3)

knitr::kable(display_table, caption = "Model Accuracy Comparison", align = 'c')
```

Based on these results, the **Radial kernel** performs best with the highest accuracy and sensitivity. This makes it the recommended choice for the Early Warning System to identify at-risk students.

Cross Validation will be elaborated in next Chapter 10.
------------------------------------------------------------------------

# 10. Model Comparison & Cross Validation

## 10.1 Model Overview
Until now, we have done the following models and experiments,
the first models served also as a exploration and learning, we did not limit ourselves to dropout and non-dropout predictions. These could also provide other insights for our client.
With NN and SVM, we focused on the dropout prediction.

<div style="font-size: 70%;">

| Model | Target Outcome | Predictors | Evaluation |
|-------|---------------|----------------|------------|
| **Linear Regression** | 2nd Sem Grade (continuous) | 1st Sem Grade, Admission Grade, GDP | R² ~ 0.68 |
| **Binomial GLM** | Graduation vs Dropout | 1st Sem Approved/Grade, Tuition Status, Debtor | AUC ~ 0.87 |
| **Poisson GLM** | 1st Sem Courses Approved (count) | Admission Grade, Previous Qualification | RMSE ~ 1.8 |
| **Gaussian GAM** | 2nd Sem Grade (continuous) | 1st Sem Grade (smooth), Age, Scholarship | R² ~ 0.72 |
| **Logistic GAM** | Graduation vs Dropout | 1st Sem Approved/Grade (smooth), Tuition, Scholarship | AUC ~ 0.89 |
| **Neural Network** | Dropout vs Not Dropout | 1st/2nd Sem Grades, Tuition, Debtor, Scholarship | Accuracy ~ 88% |
| **SVM (Linear)** | Dropout vs Not Dropout | 2nd Sem Grade, 2nd Sem Approved | Accuracy ~ 87.6% |
| **SVM (Radial)** | Dropout vs Not Dropout | 2nd Sem Grade, 2nd Sem Approved | Accuracy ~ 87.7% |
| **SVM (Poly)** | Dropout vs Not Dropout | 2nd Sem Grade, 2nd Sem Approved | Accuracy ~ 84.1% |

</div>

### Early Warning System Approaches

**Two prediction timelines are possible with our models:**

```{r early-warning-flowchart, echo=FALSE, fig.width=10, fig.height=4}
if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz("
  digraph early_warning {
    graph [rankdir=LR, bgcolor=white, fontname=Helvetica]
    node [shape=box, style='rounded,filled', fontname=Helvetica, fontsize=11]
    
    subgraph cluster_0 {
      label='Option 1: At Admission (Experimental)'
      style=dashed
      color=gray
      A [label='Admission\\nData', fillcolor='#e3f2fd']
      B [label='Poisson\\nGLM', fillcolor='#bbdefb']
      C [label='Predicted\\n1st Sem Courses', fillcolor='#90caf9']
      D [label='Binomial\\nGLM', fillcolor='#64b5f6']
      E [label='Dropout\\nRisk', fillcolor='#42a5f5']
      A -> B -> C -> D -> E
    }
    
    subgraph cluster_1 {
      label='Option 2: After 1st Semester (Recommended)'
      style=solid
      color='#2e7d32'
      F [label='1st/2nd Sem\\nData', fillcolor='#c8e6c9']
      G [label='NN or SVM', fillcolor='#81c784']
      H [label='Dropout Prediction\\n~88% Accuracy', fillcolor='#4caf50', fontcolor=white]
      F -> G -> H
    }
  }
  ")
} else {
  cat("Option 1 (Experimental): Admission Data - Poisson GLM - Predicted Courses - Binomial GLM - Dropout Risk\n")
  cat("Option 2 (Recommended): 1st/2nd Sem Data - NN or SVM - Dropout Prediction (~88% Accuracy)")
}
```

**Option 2 (NN/SVM after 1st semester)** is recommended for production use due to higher accuracy (~88%). Option 1 provides earlier predictions but relies on predicted intermediate values.

## 10.2 Cross Validation

Given the promising performance of the Support Vector Machines in Chapter 9, we proceed with a rigorous confirmation step. We use **10-Fold Cross-Validation** to test the stability of the three kernels. 

For a fair comparison, all models use the exact same configuration as the previous section:
- **Target:** Dropout vs Not Dropout
- **Predictors:** 2nd Sem Grade, 2nd Sem Approved
- **Data:** Full dataset (binary classification)

> **Note on Methodology:** To efficiently verify the stability of our *optimized* models, we use the fixed hyperparameters identified in Chapter 9 (via `tune()`) rather than re-tuning inside each fold. This validates that our chosen model configurations perform reliably across different data subsets.

```{r svm-cv-setup, echo=TRUE}
# 10-Fold Cross-Validation for SVM Kernels
set.seed(42)

# Use the 'marks' dataset from Chapter 9 (Dropout vs Not Dropout)
data_cv <- marks
n <- nrow(data_cv)

# Parameters
K <- 10
folds <- sample(rep(1:K, length.out = n))

# Storage for accuracy results
acc_linear <- numeric(K)
acc_radial <- numeric(K)
acc_poly <- numeric(K)

# Loop through k folds
for(k in 1:K) {
    # Split into train/test for this fold
    test_idx <- which(folds == k)
    train_fold <- data_cv[-test_idx, ]
    test_fold <- data_cv[test_idx, ]
    
    # Train using the TUNED parameters found in Chapter 9
    # We access 'best.parameters' from the tune objects to ensure we use the optimized values
    m_lin <- svm(y ~ x.1 + x.2, data = train_fold, kernel = "linear", 
                 cost = tune_linear$best.parameters$cost)
    
    m_rad <- svm(y ~ x.1 + x.2, data = train_fold, kernel = "radial", 
                 cost = tune_radial$best.parameters$cost, 
                 gamma = tune_radial$best.parameters$gamma)
    
    m_pol <- svm(y ~ x.1 + x.2, data = train_fold, kernel = "polynomial", 
                 cost = tune_poly$best.parameters$cost, 
                 degree = tune_poly$best.parameters$degree)
    
    # Predict & Calculate Accuracy
    p_lin <- predict(m_lin, test_fold)
    p_rad <- predict(m_rad, test_fold)
    p_pol <- predict(m_pol, test_fold)
    
    acc_linear[k] <- mean(p_lin == test_fold$y)
    acc_radial[k] <- mean(p_rad == test_fold$y)
    acc_poly[k]   <- mean(p_pol == test_fold$y)
}
```

### Cross-Validation Results

The boxplot below shows the spread of accuracy scores across the 10 folds. A narrower box indicates a more **stable** model.

```{r svm-cv-results, echo=TRUE, fig.width=8, fig.height=5}
# Combine results for plotting
cv_results <- data.frame(
  Accuracy = c(acc_linear, acc_radial, acc_poly),
  Model = factor(rep(c("Linear", "Radial", "Polynomial"), each = K))
)

# Visual comparison of stability
print(ggplot(cv_results, aes(x = Model, y = Accuracy, fill = Model)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.5) +  # Show individual fold scores
  labs(title = "Stability of SVM Kernels (10-Fold CV)",
       subtitle = "Using optimized hyperparameters from Chapter 9",
       y = "Accuracy per Fold") +
  theme_minimal() +
  scale_fill_manual(values = c("#90CAF9", "#1976D2", "#1565C0")) +
  theme(legend.position = "none"))

# Summary table
cv_summary <- cv_results %>%
  group_by(Model) %>%
  summarise(
    Mean_Accuracy = sprintf("%.1f%%", mean(Accuracy)*100),
    SD = sprintf("%.3f", sd(Accuracy)),
    Min = sprintf("%.1f%%", min(Accuracy)*100),
    Max = sprintf("%.1f%%", max(Accuracy)*100)
  )

knitr::kable(cv_summary, caption = "10-Fold CV Performance Statistics", align = 'c')
```
### Statistical Comparison

To confirm if the differences are statistically significant, we perform paired t-tests between the kernels.

```{r svm-cv-stats, echo=TRUE}
# Paired t-tests to compare kernels
t_lin_vs_rad <- t.test(acc_linear, acc_radial, paired = TRUE)
t_lin_vs_poly <- t.test(acc_linear, acc_poly, paired = TRUE)
t_rad_vs_poly <- t.test(acc_radial, acc_poly, paired = TRUE)

# Results table
stat_results <- data.frame(
  Comparison = c("Linear vs Radial", "Linear vs Poly", "Radial vs Poly"),
  p_value = c(t_lin_vs_rad$p.value, t_lin_vs_poly$p.value, t_rad_vs_poly$p.value)
)
stat_results$Significant <- stat_results$p_value < 0.05

knitr::kable(stat_results, 
             caption = "Paired t-test Results (α = 0.05)",
             digits = 4,
             align = 'c')
```

### CV Conclusion

Based on repeated 10-fold cross-validation:

*   **Radial kernel** achieves a mean accuracy of `r sprintf("%.1f%%", mean(acc_radial)*100)`.
*   **Linear kernel** achieves `r sprintf("%.1f%%", mean(acc_linear)*100)`.
*   **Polynomial kernel** achieves `r sprintf("%.1f%%", mean(acc_poly)*100)`.

This confirms the findings from Section 9: The **Radial Kernel** is the most robust and accurate model for our Early Warning System. Its ability to capture non-linear relationships gives it a crucial edge in correctly identifying at-risk students.

------------------------------------------------------------------------


# 11. Conclusions & Business Recommendations

This study applied multiple machine learning approaches to predict student dropout and graduation outcomes using academic performance data from a Portuguese higher education institution. Three classification methods were evaluated: Linear and non-linear models like GLM/GAM, Support Vector Machines with multiple kernels, and Neural Networks.

## 11.1 Key Findings
We identified **second semester academic performance** as the **strongest predictors** of student success or dropout, this was proven in our most reliable models SVM and NN features selection process.
However, for the business case, first semester **early warning** could make more sense, as first semester performance comes as second often as the most important features.

*   **1st Semester Performance:** The critical **"early warning"** signal. While less definitive than 2nd-semester results, poor performance here is the earliest actionable indicator of risk.
*   **2nd Semester Performance:** The **"definitive"** predictor. Models using 2nd-semester data (SVM, NN) achieved the highest accuracy (~88%). By the end of the first year, the student's success or dropout is largely decided. **Failure to pass enrolled units** in the second semester is the dominant risk factor.

The **early warning system** could be established by the best-performing models **SVM** or **Neural Network**. The accuracy rate could reach **86.8%**. Another **experimental system** could be  [Admission Data] - [Poisson GLM] - [Predicted 1st Sem Courses] - [Binomial GLM] - [Dropout Risk].

## 11.2 Business Recommendations

Our models suggest that **grade performance** and **financial stability** are the most critical factors for student retention. We propose the following structured intervention strategy:

**1. For Our Client University (Two-Stage Filter):**

*   **Implement "Option 2" Warning System:**
*   **Stage 1 (After Sem 1):** Deploy the **Binomial GLM** or **1st-Sem SVM** as a broad screening tool. The goal is *prevention*, flag students with difficulty early.
*   **Stage 2 (After Sem 2):** Deploy the higher precision **Neural Network** or **SVM** . The goal is *crisis intervention*. Accurately identify students who are on the brink of dropout for intensive counseling.
*   **Human-in-the-Loop Protocol:** Use the model as a *filtering tool*. High-risk lists should be reviewed by advisors, not used for automated decisions.

**2. For Academic Advisors:**
*   **Early Warning is key:** Do not wait for the 2nd semester. If the **system predicts dropout** based on first-semester data (where failing multiple courses is a key driver), this is the single biggest red flag for action. And another drop-out signal in 2nd semester would be definitive signal for immediate intervention.
*   **Multi-faceted Support:** Combine academic tutoring with financial counseling, as financial stress (debtor status) is a key dropout factor too.

**3. For Policy Makers:**
*   **Economic Context:** External factors like Unemployment Rate and GDP have minor effects on academic success (The Linear Regression coefficients (Section 4.8) show GDP/Unemployment have small effect sizes). This could be considered into policy making as background information.
*   **Financial Safety Nets:** Consider policy to provide scholarship and emergency loan programs for in-risk students.

# 12. Limitations & Future Work

## 12.1 Limitations

Our dataset focuses on academic records and financial indicators, but real student life is more complex. We don't have data on **motivation**, **mental health**, or **study habits** (like how often someone visits the library or attends tutoring). These "soft" factors likely matter a lot, but they're hard to collect.

Finally, this is a Portuguese institution with its own grading scale (0-20). What works here might not directly transfer to HSLU or other universities without recalibration. A standardized scoring system would be needed for broader use.

## 12.2 Ideas for Future Work

1. **Behavioral Signals:** Library check-ins, Ilias login frequency, and assignment submission times could significantly boost predictive power.
2. **Cross-University Validation:** Scaling this system to other universities (e.g., HSLU) to test robustness across different institutions.

# 13. References

1. OECD (2025). *Education at a Glance 2025*. OECD. https://www.oecd.org/en/publications/education-at-a-glance-2025_1a3543e2-en/switzerland_e7c7fc03-en.html
2. Realinho, V., Machado, J., Baptista, L., & Martins, M.V. (2022). *Predict Students' Dropout and Academic Success*. UCI Machine Learning Repository. https://doi.org/10.24432/C5MC89

-------------------------------------------------------------------------

# 14. Generative AI Declaration & Guidelines

Generative AI was used as a supplementary tool on top of the assisting material provided in the Machine Learning1 lecture.

## Guidelines for Responsible and Effective Usage of Gen-AI
1. **Human-in-the-loop verification:** All AI-suggested content was treated as a draft. Text, code, or sources delivered by AI had to be critically reviewed and tested before inclusion.
2. **Emphasis on learning:** Gen-AI functioned as a learning companion rather than an automated code generator. After receiving suggestions, team members engaged in follow-up questions to grasp the underlying logic and method.
3. **Transparency:** The team openly acknowledged where and how AI was used in the workflow so the benefits and quality safeguards remained clear.

## Use Cases of AI Tools in the Machine Learning 1 Course Context
AI-based assistants were applied in the following specific cases:

1. **Brainstorming:** Elaborating initial ideas, structuring thoughts, and outlining coding approaches.
2. **Debugging & optimization:** Explaining error messages and helping improve self-developed scripts to raise efficiency.
3. **Proofreading:** Accelerating grammar and typo checks during documentation.
4. **Information acquisition:** Searching for methodological references or code documentation for R Markdown, or machine learning methods (always followed by human verification of credibility).

## Benefits and Challenges in Using Generative AI Tools

1. **Benefits:** Faster debugging (e.g., resolving R Markdown knitting errors or debugging issues), on-demand tutoring for advanced questions, and more time spent on analytical reasoning instead of routine fixes.
2. **Challenges:** The ease of getting answers can create a temptation to trust outputs blindly. The "human-in-the-loop" guideline ensured the team retained ownership, validating every AI-assisted contribution.

-------------------------------------------------------------------------