---
title: "Cereal Ratings using a Neural Network"
author: "Daniel Meister"
date: "10/23/2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Adapted from _https://www.analyticsvidhya.com/blog/2017/09/creating-visualizing-neural-network-in-r/_

## Load Packages

```{r}
library(tidyverse)
library(caret)
library(neuralnet)
```


## Load the data

```{r}
cereals <- read_csv("cereals.csv")
```

```{r}
str(cereals)
```


## Prepare for Training

```{r}
set.seed(123)
indices <- createDataPartition(cereals$rating, p = 0.8, list = FALSE)
train <- cereals %>% slice(indices)
test <- cereals %>% slice(-indices)
boxplot(train$rating, test$rating, cereals %>% sample_frac(0.2) %>% pull(rating))
```

And this time we will also scale the data (by hand for educational purposes)

```{r}
max <- apply(cereals, 2, max)
min <- apply(cereals, 2, min)
cereals_scaled <- as.data.frame(scale(cereals, center = min, scale = max - min))
train_scaled <- cereals_scaled %>% slice(indices)
test_scaled <- cereals_scaled %>% slice(-indices)
```

## Fit the Network

```{r}
set.seed(42)
cereal_net = neuralnet(rating ~ calories + protein + fat + sodium + fiber, train_scaled, hidden = 3 , linear.output = TRUE)
plot(cereal_net)
```
## Predict the test set

```{r}
pred <- compute(cereal_net, test %>% select(-rating))
pred$net.result
```
So what went wrong here? Obviously we also need to use the scaled data to make predictions and then scale back to real rating values

```{r}
pred_scaled <- compute(cereal_net, test_scaled %>% select(-rating))
pred <- pred_scaled$net.result * (max(cereals$rating) - min(cereals$rating)) + min(cereals$rating)
pred
```


```{r}
plot(test$rating, pred, col='blue', pch=16, ylab = "predicted rating NN", xlab = "real rating")
abline(0,1)
```
And calculate the RMSE
```{r}
sqrt(mean((test$rating - pred)^2))
```

## Redo with `caret` and Cross Validation

In principle we cannot be sure that we were not simply "lucky" with the train/test split above, so the proper way to run this would be via `caret` using k-fold Cross Validation. And also as we use `caret` anyway let's also run some hyperparameter tuning:

```{r message=FALSE, warning=FALSE}
set.seed(42)
tuGrid <- expand.grid(.layer1=c(1:4), .layer2=c(0,2), .layer3=c(0))

trCtrl <- trainControl(
  method = 'repeatedcv', 
  number = 5, 
  repeats = 10, 
  returnResamp = 'final'
)

models <- train(
  x = cereals %>% select(-rating),
  y = cereals_scaled %>% pull(rating),
  method = 'neuralnet', metric = 'RMSE', 
  linear.output = TRUE,
  # be careful, does only work on x!
  preProcess = c('center', 'scale'),
  tuneGrid = tuGrid,
  trControl = trCtrl
)
```

```{r}
plot(models)
```
and extract the best model

```{r}
plot(models$finalModel)
```

