---
title: "Support Vector Machine Analysis - Student Dropout Prediction"
subtitle: "Understanding the Pathways to Academic Achievement"
date: "`r Sys.Date()`"
output: # output both html and pdf, need to check if both output aligns in the end
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,           # Show code by default (can be hidden with code_folding in HTML)
  message = FALSE,       # Hide package loading messages
  warning = FALSE,       # Hide warnings
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

# Load required libraries
library(tidyverse)
library(ggplot2)
library(e1071)
library(caret)
```

# Support Vector Machine Analysis

We will now test different vector models to understand what is the best classifier for our student's performance. To classify our dataset, the model plots a separating line or boundary between the two outcome classes using a number of predictors - in order to keep it clear, we want to create a two-dimensional plane (two predictors).

By applying different kernels—linear, radial, and polynomial—we assess how the shape of the decision boundary affects prediction accuracy and determine which kernel provides the most reliable classification for our dataset.

## Feature Selection

For us to choose the best predictors, we will compute F-statistic ANOVA for every numeric feature in the dataset. The F-score tell us how much the mean of each feature differs between the two target groups (graduated vs. not graduated). higher scores mean there is a bigger separation between classes, and a better cutoff between groups. 
Results indicate that grades and the amount of credits approved have the highest F-Values, suggesting they are the most significant predictors for predicting the success of graduating or not.

In this case, we want to choose the features of second semester marks and the amount of units passed to plot our model.

```{r fscore-top5}
numeric_cols <- sapply(data, is.numeric)
features <- names(data)[numeric_cols]

results <- data.frame(
  Feature = character(),
  F_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (feat in features) {
  formula <- as.formula(paste(feat, "~ Target"))
  anova_result <- aov(formula, data = data)
  summary_anova <- summary(anova_result)
  
  results <- rbind(
    results,
    data.frame(
      Feature = feat,
      F_value = round(summary_anova[[1]]$`F value`[1], 2),
      p_value = summary_anova[[1]]$`Pr(>F)`[1]
    )
  )
}

results <- results[order(-results$F_value), ]
head(results, 5)


## Exploratory Visualization


## Train/Test Split

Now, we will split the data into training and testing sets, and then fit several SVM models to classify the students based on their marks.

```{r train-test-split}
set.seed(123)
trainIndex <- createDataPartition(marks$y, p = 0.8, list = FALSE)
train_data <- marks[trainIndex, ]
test_data <- marks[-trainIndex, ]

cat("Training set size:", nrow(train_data), "\n")
cat("Test set size:", nrow(test_data), "\n")
```

## Model Training

Now, we will fit the SVM model with a linear, a radial and a polynomial kernel.

```{r train-models}
svm_linear <- svm(y ~ x.1 + x.2, data = train_data, kernel = "linear", cost = 1)
svm_radial <- svm(y ~ x.1 + x.2, data = train_data, kernel = "radial", cost = 10, gamma = 0.1)
svm_poly <- svm(y ~ x.1 + x.2, data = train_data, kernel = "polynomial", cost = 1, degree = 3)
```

## Model Evaluation

Predict and compare the models

```{r predictions}
pred_linear <- predict(svm_linear, test_data)
pred_radial <- predict(svm_radial, test_data)
pred_poly <- predict(svm_poly, test_data)
```

### Linear Kernel Results

```{r eval-linear}
cat("\n Linear Kernel Result:\n")
cm_linear <- confusionMatrix(pred_linear, test_data$y)
print(cm_linear)
```

### Radial Kernel Results

```{r eval-radial}
cat("\n Radial Kernel Result:\n")
cm_radial <- confusionMatrix(pred_radial, test_data$y)
print(cm_radial)
```

### Polynomial Kernel Results

```{r eval-poly}
cat("\n Polynomial Kernel Result:\n")
cm_poly <- confusionMatrix(pred_poly, test_data$y)
print(cm_poly)
```

## Performance Comparison

```{r comparison}
results <- data.frame(
  Model = c("Linear", "Radial", "Polynomial"),
  Accuracy = c(cm_linear$overall['Accuracy'],
               cm_radial$overall['Accuracy'],
               cm_poly$overall['Accuracy'])
)

knitr::kable(results, digits = 3)
```

## Decision Boundary Visualizations

### Linear Kernel

```{r plot-linear, fig.width=8, fig.height=6}
plot(svm_linear, train_data, x.1 ~ x.2,
     svSymbol = 1, dataSymbol = 16,
     symbolPalette = c("#ff2000", "#828282", "#2ecc71"),
     color.palette = terrain.colors)
title("Linear Kernel SVM")
```

### Radial Kernel

```{r plot-radial, fig.width=8, fig.height=6}
plot(svm_radial, train_data, x.1 ~ x.2,
     svSymbol = 1, dataSymbol = 16,
     symbolPalette = c("#ff2000", "#828282", "#2ecc71"),
     color.palette = terrain.colors)
title("Radial Kernel SVM")
```

### Polynomial Kernel

```{r plot-poly, fig.width=8, fig.height=6}
plot(svm_poly, train_data, x.1 ~ x.2,
     svSymbol = 1, dataSymbol = 16,
     symbolPalette = c("#ff2000", "#828282", "#2ecc71"),
     color.palette = terrain.colors)
title("Polynomial Kernel SVM")
```
