---
title: "Iris Data Classification using a Support Vector Machine"
author: "Daniel Meister"
date: "10/16/2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the Packages

First make sure we have all the functionality from the `tidyverse` available:

```{r}
library(tidyverse)
theme_set(theme_bw())
```

We are going to use the `e1071` package for this example and also some helper functions from `caret`

```{r}
library(e1071) 
library(caret)
```
## Loading the Data

Load the `iris` data set into the environment

```{r}
data(iris)
```

and let's have a look a the structure.

```{r}
str(iris)
```
## Have a quick look at the Data

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()
```
```{r}
iris %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point()
```

## Prepare the Data for Training

```{r}
set.seed(123)
indices <- createDataPartition(iris$Species, p=.85, list = F)
```

### Create some easy Variables to access Data

```{r}
train <- iris %>%
  slice(indices)
test_in <- iris %>%
  slice(-indices) %>%
  select(-Species)
test_truth <- iris %>%
  slice(-indices) %>%
  pull(Species)
```

## Train the Support Vector Machine

Call the `svm` function using the default `cost = 10` parameter.

```{r}
set.seed(123)
iris_svm <- svm(Species ~ ., train, kernel = "linear", scale = TRUE, cost = 10)
```

Of course the visualization is a bit difficult as we have some classification regions in four dimensions.

So plot first the resulting classification for the two Petal leaf dimensions

```{r}
plot(iris_svm, train, Petal.Length ~ Petal.Width)
```
And for the Sepal leaf dimensions (of course you need to slice the other dimensions at a reasonable point)


```{r}
plot(iris_svm, train, Sepal.Length ~ Sepal.Width, slice = list(Petal.Length = 4.5, Petal.Width = 1.75))
```

## Make Predictions

```{r}
test_pred <- predict(iris_svm, test_in)
table(test_pred)
```
## Evaluate the Results

```{r}
conf_matrix <- confusionMatrix(test_pred, test_truth)
conf_matrix
```
So we have a resulting SVM with a prediction accuracy of around 95% (of course we would also need to do some cross validation to ensure we were not simply "lucky" with the train/test split).

## Try again with other parameters


```{r}
set.seed(123)
iris_svm2 <- svm(Species ~ ., train, kernel = "radial", scale = TRUE, cost = 100)
summary(iris_svm2)
```

```{r}
plot(iris_svm2, train, Petal.Length ~ Petal.Width, slice = list(Sepal.Length = 6, Sepal.Width = 3))
```



```{r}
plot(iris_svm2, train, Sepal.Length ~ Sepal.Width, slice = list(Petal.Length = 4.5, Petal.Width = 1.75))
```
```{r}
test_pred <- predict(iris_svm2, test_in)
table(test_pred)
```
```{r}
conf_matrix <- confusionMatrix(test_pred, test_truth)
conf_matrix
```


So we have here the same accuracy but with much more complex boundaries (radial kernels and higher costs) so we might be overfitting here a bit more.