---
title: "Neural Network Analysis - Student Dropout Prediction"
subtitle: "Understanding the Pathways to Academic Achievement"
date: "`r Sys.Date()`"
output: # output both html and pdf, need to check if both output aligns in the end
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,           # Show code by default (can be hidden with code_folding in HTML)
  message = FALSE,       # Hide package loading messages
  warning = FALSE,       # Hide warnings
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

# Load required libraries
library(tidyverse)
library(caret)
library(nnet)         # Lab 1 style neural network
library(neuralnet)    # Lab 2 style neural network
library(ROCR)
library(NeuralNetTools)  # Visualize network structure and importances
library(gridExtra)    # For side-by-side plots

```
##  next steps: 0. change template format for code chunks, how it is shown for everyones template 1. need to finish lab2 neuralnet 2. need to interprete the results 3. need to refine the format and analysis(code chunks how shown etc) check length 4. need to compare with lab1 lab2 and move to notes* 
## Objective
We want our neural network to play the role of an early-warning whether a student is likely to drop out or graduate by learning the messy, non-linear patterns hidden in data.

## Feature Selection

We focus on three feature blocks from the preprocessed dataset:

1. **First semesters performance:** 1st- and 2nd-semester grades plus approved credits capture the strongest signals from our ANOVA screening during data preprocessing.
2. **Financial pressure/support:** Tuition fees up to date, debtor status, and scholarship holder tell us whether financial stress the student faces.
3. **Entry profile:** Admission grade and age at enrollment anchor the model with information about students’ starting point.

## Data Loading & Preprocessing

```{r load-data}
# Note: When pasting into final_report_ml1_group.rmd, remove this line (data already loaded)
data <- read.csv("../../data/preprocessed_data.csv", stringsAsFactors = TRUE)

# Keep all three outcome classes so NN addresses the same business question
data$Target <- factor(data$Target, levels = c("Dropout", "Enrolled", "Graduate"))

# Selected features
nn_features <- c(
  "Curricular.units.1st.sem..approved.",
  "Curricular.units.1st.sem..grade.",
  "Curricular.units.2nd.sem..approved.",
  "Curricular.units.2nd.sem..grade.",
  "Tuition.fees.up.to.date",
  "Scholarship.holder",
  "Age.at.enrollment",
  "Admission.grade",
  "Debtor"
)

# Create modeling dataset
nn_model_data <- data %>%
  select(all_of(nn_features), Target) %>%
  na.omit()

# Scaling (important for neural networks)
numeric_cols <- sapply(nn_model_data, is.numeric)
nn_model_data[, numeric_cols] <- scale(nn_model_data[, numeric_cols])

cat("Dataset size:", nrow(nn_model_data), "students\n") # cat() is R function that prints text to console in  single line
cat("Class distribution:\n")
table(nn_model_data$Target) # counts how many records fall into each target class to preview
```

## Train/Test Split

```{r train-test-split}
set.seed(123)
train_idx <- createDataPartition(nn_model_data$Target, p = 0.8, list = FALSE) # not using runif(nrow(GermanCredit)) < 0.8 simply to sample ~80% of rows without worrying about class balance
train_nn <- nn_model_data[train_idx, ]
test_nn <- nn_model_data[-train_idx, ]

cat("Training set:", nrow(train_nn), "samples\n")
cat("Test set:", nrow(test_nn), "samples\n")
```

## Model Training

We train two neural networks to compare Lab 1 and Lab 2 approaches.

### Model A: Single Hidden Layer (Lab 1 - nnet)

Following `ANN_Lab1_GermanCredit_v1.Rmd`, we use `nnet` with one hidden layer and weight decay.

```{r train-model-A}
# Model A: nnet with single hidden layer (Lab 1 style)
model_A <- nnet(
  Target ~ .,
  data = train_nn,
  size = 10,        # 10 hidden units
  decay = 0.01,     # Regularization
  maxit = 500,      # Maximum iterations
  trace = FALSE     # Don't print training progress
)
```

### Model B: Three Hidden Layers (Lab 2 - neuralnet)

Following `ANN_Lab2_Iris_v1.Rmd`, we use `neuralnet` with a deeper (three-layer) architecture. This requires one-hot encoding since `neuralnet` needs numeric inputs only.

```{r train-model-B}
# One-hot encode predictors for neuralnet (Lab 2 approach)
dummies <- dummyVars(Target ~ ., data = train_nn)
train_matrix <- data.frame(predict(dummies, newdata = train_nn))
test_matrix <- data.frame(predict(dummies, newdata = test_nn))

# Create dummy columns for 3-class target (neuralnet needs numeric output)
train_matrix$Target_Dropout <- as.numeric(train_nn$Target == "Dropout")
train_matrix$Target_Enrolled <- as.numeric(train_nn$Target == "Enrolled")
train_matrix$Target_Graduate <- as.numeric(train_nn$Target == "Graduate")

# Build formula: output columns ~ input columns
input_names <- names(train_matrix)[!grepl("Target", names(train_matrix))]
output_names <- c("Target_Dropout", "Target_Enrolled", "Target_Graduate")
formula_B <- as.formula(paste(
  paste(output_names, collapse = " + "), "~",
  paste(input_names, collapse = " + ")
))

# Model B: neuralnet with three hidden layers (Lab 2 style)
model_B <- neuralnet(
  formula_B,
  data = train_matrix,
  hidden = c(5,7,5),   # Three hidden layers: 5 neurons, then 7 and 5 neurons
  linear.output = FALSE,
  threshold = 0.05,
  stepmax = 1e6        # big number to allow more iterations
)
```

## Model Evaluation

We generate predictions and confusion matrices for both models.

```{r evaluate-both-models}
# --- Model A predictions (nnet) ---
pred_A_class <- predict(model_A, newdata = test_nn, type = "class")
conf_matrix_A <- confusionMatrix(as.factor(pred_A_class), test_nn$Target)

# --- Model B predictions (neuralnet) ---
# compute() runs the network on new data and returns raw output values
pred_B_raw <- compute(model_B, test_matrix[, input_names])$net.result
# Pick the class with highest probability for each student
pred_B_class <- factor(
  c("Dropout", "Enrolled", "Graduate")[apply(pred_B_raw, 1, which.max)],
  levels = c("Dropout", "Enrolled", "Graduate")
)
conf_matrix_B <- confusionMatrix(pred_B_class, test_nn$Target)
```

## Performance Metrics Summary

We report macro-averaged metrics for both models to compare their performance.

```{r performance-summary}
cat("=== Model A (nnet - 1 hidden layer) ===\n")
cat("Accuracy:", round(conf_matrix_A$overall["Accuracy"], 3), "\n")
by_class_A <- conf_matrix_A$byClass
cat("Precision (macro):", round(mean(by_class_A[,"Precision"], na.rm = TRUE), 3), "\n")
cat("Recall (macro):", round(mean(by_class_A[,"Recall"], na.rm = TRUE), 3), "\n\n")

cat("=== Model B (neuralnet - 3 hidden layers) ===\n")
cat("Accuracy:", round(conf_matrix_B$overall["Accuracy"], 3), "\n")
by_class_B <- conf_matrix_B$byClass
cat("Precision (macro):", round(mean(by_class_B[,"Precision"], na.rm = TRUE), 3), "\n")
cat("Recall (macro):", round(mean(by_class_B[,"Recall"], na.rm = TRUE), 3), "\n")
```

Model A (1 layer) still delivers the strongest result on the test set (Accuracy 0.764, Macro Precision 0.703, Macro Recall 0.666). The deeper Model B (three layers: 5-7-5) improved slightly over the earlier two-layer trial but remains lower overall (Accuracy 0.747, Macro Precision 0.681, Macro Recall 0.652). This mirrors the lecture message that more layers are not automatically better for structured tabular data.

## Confusion Matrix Comparison

Side-by-side heatmaps comparing both models. X-axis (top) = Predicted, Y-axis = True labels.

```{r confusion-matrix-comparison, fig.width=12, fig.height=5}
# Helper function to create confusion matrix plot
# Rows = True values (y-axis), Columns = Predicted values (x-axis on top)
plot_confusion_matrix <- function(conf_matrix, title) {
  cm_df <- as.data.frame(conf_matrix$table)
  names(cm_df) <- c("Predicted", "True", "Count")
  
  ggplot(cm_df, aes(x = Predicted, y = True, fill = Count)) +
    geom_tile(color = "white", linewidth = 1) +
    geom_text(aes(label = Count), size = 5, fontface = "bold", color = "white") +
    scale_fill_gradient(low = "#3498DB", high = "#E74C3C", name = "Count") +
    scale_x_discrete(position = "top") +
    scale_y_discrete(limits = rev(c("Dropout", "Enrolled", "Graduate"))) +
    labs(title = title, x = "Predicted Class", y = "True Class") +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
      axis.text = element_text(size = 10),
      legend.position = "right"
    )
}

# Create plots for both models
plot_A <- plot_confusion_matrix(conf_matrix_A, "Model A: nnet (1 hidden layer)")
plot_B <- plot_confusion_matrix(conf_matrix_B, "Model B: neuralnet (3 hidden layers)")

# Display side by side
grid.arrange(plot_A, plot_B, ncol = 2)
```

## ROC Curve Comparison

We compare ROC curves for the "Graduate" class (one-vs-rest) across both models using `ROCR`.

```{r roc-comparison, fig.width=12, fig.height=5}
# --- Model A ROC (nnet) ---
prob_A <- predict(model_A, newdata = test_nn, type = "raw")
pred_A_grad <- ROCR::prediction(prob_A[, "Graduate"], test_nn$Target == "Graduate")
perf_A_grad <- ROCR::performance(pred_A_grad, "tpr", "fpr")
auc_A <- round(ROCR::performance(pred_A_grad, "auc")@y.values[[1]], 3)

# --- Model B ROC (neuralnet) ---
# Graduate is the 3rd output column in pred_B_raw
prob_B <- pred_B_raw[, 3]
pred_B_grad <- ROCR::prediction(prob_B, test_nn$Target == "Graduate")
perf_B_grad <- ROCR::performance(pred_B_grad, "tpr", "fpr")
auc_B <- round(ROCR::performance(pred_B_grad, "auc")@y.values[[1]], 3)

# Print AUC values
cat("AUC for Graduate vs. Rest:\n")
cat("  Model A (nnet):", auc_A, "\n")
cat("  Model B (neuralnet):", auc_B, "\n")

# Side-by-side ROC plots
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

# Model A ROC
plot(perf_A_grad, col = "#2ecc71", lwd = 2, main = "Model A: nnet (1 hidden layer)")
abline(a = 0, b = 1, lty = 2, col = "gray")
legend("bottomright", legend = paste("AUC =", auc_A), bty = "n", cex = 1.1)

# Model B ROC
plot(perf_B_grad, col = "#3498db", lwd = 2, main = "Model B: neuralnet (3 hidden layers)")
abline(a = 0, b = 1, lty = 2, col = "gray")
legend("bottomright", legend = paste("AUC =", auc_B), bty = "n", cex = 1.1)

par(mfrow = c(1, 1))  # Reset layout
```

## Feature Importance (Permutation)

Permutation importance (same spirit as Lab 1’s “what-if” checks) measures how much each model’s accuracy drops when we randomly shuffle one predictor at a time. Bigger drops mean the learner relied heavily on that feature.

```{r permutation-importance, message=FALSE}
# Helper to compute accuracy drop for a single feature and model
permute_feature <- function(model_type, var_name) {
  permuted <- test_nn
  permuted[[var_name]] <- sample(permuted[[var_name]])
  
  if (model_type == "Model A: nnet (1 layer)") {
    baseline <- mean(predict(model_A, newdata = test_nn, type = "class") == test_nn$Target)
    perm_acc <- mean(predict(model_A, newdata = permuted, type = "class") == test_nn$Target)
  } else {
    # Rebuild dummy matrix for permuted data
    permuted_matrix <- data.frame(predict(dummies, newdata = permuted))
    class_map <- c("Dropout", "Enrolled", "Graduate")
    baseline <- mean(pred_B_class == test_nn$Target)
    perm_raw <- compute(model_B, permuted_matrix[, input_names])$net.result
    perm_class <- factor(class_map[apply(perm_raw, 1, which.max)], levels = class_map)
    perm_acc <- mean(perm_class == test_nn$Target)
  }
  
  tibble(
    Model = model_type,
    Feature = var_name,
    AccuracyDrop = baseline - perm_acc
  )
}

perm_results <- map_dfr(
  nn_features,
  ~bind_rows(
    permute_feature("Model A: nnet (1 layer)", .x),
    permute_feature("Model B: neuralnet (3 layers)", .x)
  )
)

# Friendlier labels for the plot
feature_labels <- c(
  "Curricular.units.1st.sem..approved." = "1st Sem Credits Approved",
  "Curricular.units.1st.sem..grade." = "1st Sem Grade",
  "Curricular.units.2nd.sem..approved." = "2nd Sem Credits Approved",
  "Curricular.units.2nd.sem..grade." = "2nd Sem Grade",
  "Tuition.fees.up.to.date" = "Tuition Fees Up to Date",
  "Scholarship.holder" = "Scholarship Holder",
  "Age.at.enrollment" = "Age at Enrollment",
  "Admission.grade" = "Admission Grade",
  "Debtor" = "Debtor Status"
)

perm_results <- perm_results %>%
  mutate(Feature = recode(Feature, !!!feature_labels)) %>%
  group_by(Model) %>%
  mutate(Feature = fct_reorder(Feature, AccuracyDrop)) %>%
  ungroup()

ggplot(perm_results, aes(x = AccuracyDrop, y = Feature, fill = Model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Model, scales = "free_y") +
  labs(
    title = "Permutation Importance (Accuracy Drop when Shuffled)",
    x = "Accuracy Drop",
    y = "Feature"
  ) +
  scale_fill_manual(values = c("#5DADE2", "#48C9B0")) +
  theme_minimal()
```

## Neural Network Architecture Visualization

`NeuralNetTools::plotnet()` turns the fitted networks into diagrams that show layers, neurons, and connection strengths (thicker lines = larger absolute weights). Seeing both side by side links back to Lab 2’s focus on architecture design.

```{r nn-visualization, fig.width=12, fig.height=6}
input_labels <- recode(nn_features, !!!feature_labels)
output_labels <- c("Dropout", "Enrolled", "Graduate")

par(mfrow = c(1, 2), mar = c(0, 8, 2, 6))  # widen margins so labels fit
NeuralNetTools::plotnet(model_A,
        alpha = 0.7,
        circle_col = list("steelblue", "skyblue"),
        x_names = input_labels,
        y_names = output_labels,
        max_sp = TRUE,
        main = "Model A: nnet (1 hidden layer)")
NeuralNetTools::plotnet(model_B,
        alpha = 0.7,
        circle_col = list("darkseagreen", "palegreen"),
        x_names = input_labels,
        y_names = output_labels,
        max_sp = TRUE,
        main = "Model B: neuralnet (3 hidden layers)")
par(mfrow = c(1, 1))
```

## Findings & Interpretation

**Model A (nnet):** 9 inputs → 10 hidden neurons → 3 outputs. Simple, fast, handles factors directly and reaches Accuracy 0.764 / AUC 0.907.

**Model B (neuralnet):** 9 inputs → 5 → 7 → 5 neurons → 3 outputs. Shows Lab 2’s deeper setup but, even after extending to three layers, levels off at Accuracy 0.747 / AUC 0.887. The confusion matrix heatmap highlights slightly more mistakes on Dropout and Graduate rows, and the ROC curve is a bit flatter than Model A’s.

**Strengths observed:** Neural networks capture the non-linear interaction between grades, tuition status, and scholarships. Model A identifies Graduate students especially well (408 correct) while still catching most Dropout cases. Permutation importance confirms that 1st- and 2nd-semester performance plus tuition status drive most of the predictive power.

**Limitations observed:**

- Deeper networks remain sensitive to hyperparameters and can overfit small samples.
- Interpretability is limited compared to GLM/GAM coefficients.
- Enrolled students are the hardest class for both models (recall below 0.30), suggesting we may need additional features.

## Conclusions

1. Both neural networks classify student outcomes reasonably well, but the single hidden layer provides the best balance of accuracy and stability on this dataset.
2. Adding more layers (5-7-5) slightly improves over the earlier 2-layer experiment yet still trails the simpler architecture; this reinforces the lab insight that “deeper” is not automatically “better” for structured data.
3. Standard preprocessing steps from Lab 2 (scaling, stratified split) remain essential for consistent results.

**Recommendation:** if not enough space in report, Keep Model A as the primary neural network in this case, and mention Model B as an exploratory Lab 2 extension